---
title: "Basic Statistics and Sampling"
author: "Jerome Dumortier"
date: "`r format(Sys.time(),'%d %B %Y')`"
output: beamer_presentation
---

## Lecture Overview
Topics covered:

- Law of Large Numbers
- Central Limit Theorem

## Law of Large Numbers
Measuring unemployment rate in the United States:

- Current Population Survey (CPS)
- Monthly survey among 60,000 households
- Classification: *Employed*, *Unemployed*, *Not in the labor force*

Law of large numbers:

- Any feature of a distribution can be recovered from repeated sampling.

Example of flipping a coin:

- Two possible outcomes: Heads or tails
- Key condition: Independence
- Expected value of heads (or tails): $E(H)=E(T)=0.5$

Difficulty to predict the share of heads from a single coin flip but high prediction precision from several thousand flips.

## Law of Large Numbers: Flipping a Coin
```{r, echo=FALSE,warning=FALSE}
library(ggplot2)
n   = 1000
x   = sample(0:1,n,repl=T)
s   = cumsum(x)
r   = s/(1:n)
lln = data.frame(trials=c(1:1000),heads=r)
ggplot(lln,aes(x=trials,y=heads))+geom_line()+theme_bw()+ylab("Share of Heads")+ylim(0,1)+xlab("Trials")
rm(n,x,s,r,lln)
```

## Refresher: Sample versus Population
Why sampling is necessary:

- Sampling the entire population may be expensive or impossible.
- Sampling the entire population may be destructive (e.g., sampling all tires).

Random sample:

- Every item or person in the population (more specifically sample frame) has the same probability of getting selected into the sample.

Example for polling before an election:

- Every person with voting rights is in the sample frame and has the same chance of getting selected by a news agency for polling.

## Estimation of the Sample Mean and the Sample Variance
Estimation of the population mean based on a sample:
$$\bar{x} = \frac{1}{N} \sum^{N}_{i=1} x_i$$
Estimation of the population variance based on a sample:
$$s^2 = \frac{1}{N-1} \sum^{N}_{i=1} (x_i-\bar{x})^2$$
And this is important:

- In R, `var()` and `sd()` calculate the variance assuming a sample, i.e., division by $N-1$.

## Illustration: Estimating the Population Variance I
What we know about the population:

- Population size: 100,000
- Mean: $\mu=50$
- Standard deviation: $\sigma=20$

Sampling:

- Sample size ranging from 2 to 50
- Repeating the sampling 1000 times

## Illustration: Estimating the Population Variance II
```{r,echo=FALSE}
pop       = rnorm(100000,mean=50,sd=20)
n_nminus1 = matrix(0,1000,2)
out       = matrix(0,50,2)
for (j in 2:50){
     for (i in 1:1000){
          temp            = sample(pop,j)
          n_nminus1[i,2]  = var(temp)
          n_nminus1[i,1]  = var(temp)/j*(j-1)}
     out[j,1] = mean(n_nminus1[,1])
     out[j,2] = mean(n_nminus1[,2])}
out                 = sqrt(out[2:50,])
out1                = data.frame(Sample=c(2:50),Estimate=out[,1],Method="Dividing by N")
out2                = data.frame(Sample=c(2:50),Estimate=out[,2],Method="Dividing by N-1")
out                 = rbind(out1,out2)
ggplot(out,aes(x=Sample,y=Estimate,color=Method))+geom_line(aes(color=Method))+theme_bw()+
     ylim(0,25)+theme(legend.position="bottom")
rm(n_nminus1,out,out1,out2,i,j,pop,temp)
```


## Sampling Distribution and Central Limit Theorem
A statistic is a random variable (with its own probability distribution) based on a sample. For example, repeated polling of 1,000 people about their political preferences will result in a different outcome each time. For the sampling distribution of the mean $\bar{x}$, we have the following:

- Mean of the sampling distribution: $\mu_{\bar{X}}$
- Variance of the sampling distribution: $\sigma^2_{\bar{X}}$
- Standard deviation of the sampling distribution (commonly known as standard error): $\sigma_{\bar{X}}$

Central Limit Theorem

- Independent of the underlying distribution, as the sample size increases, the sampling distribution of the mean will follow a normal distribution.

## Central Limit Theorem: Illustration
```{r,echo=FALSE}
pop_n               = 10000
sample_n            = 500
pop_uniform         = runif(pop_n,min=0,max=1)
pop_poisson         = rpois(pop_n,lambda=5)
pop_exponential     = rexp(pop_n,rate=1.5)
pop_beta            = rbeta(pop_n,0.5,0.5)
out                 = matrix(0,1000,4);
for (i in 1:1000){
     out[i,1]       = mean(sample(pop_uniform,sample_n))
     out[i,2]       = mean(sample(pop_poisson,sample_n))
     out[i,3]       = mean(sample(pop_exponential,sample_n))
     out[i,4]       = mean(sample(pop_beta,sample_n))}
  par(mfrow=c(2,4))
    hist(pop_uniform,main="Population: Uniform",xlab="")
    hist(out[,1],main="Sample: Uniform",xlab="")
    hist(pop_poisson,main="Population: Poisson",xlab="")
    hist(out[,2],main="sample: Poisson",xlab="")
    hist(pop_exponential,main="Population: Exponential",xlab="")
    hist(out[,3],main="Sample: Exponential",xlab="")
    hist(pop_exponential,main="Population: Beta",xlab="")
    hist(out[,4],main="Sample: Beta",xlab="")
```

## Central Limit Theorem: Implications for Estimation
The standard error of the mean is given by:
$$\sigma_{\bar{x}} = \sqrt{\frac{\sigma^2}{n}} =  \frac{\sigma}{\sqrt{n}}$$
The sample standard deviation is the statistic defined by:
$$s=\sqrt{s^2}$$
Suppose you have to predict the share of heads after flipping a coin multiple times. The variance of $n$ coin flips is:
$$\text{Var}(n) = \frac{p \cdot (1-p)}{n}$$

Hence: $\text{Var}(1) = 0.5$, $\text{Var}(10) = 0.025$, $\text{Var}(1000) = 0.00025$, etc.

## Application: Insurance Market
Risk aversion for individuals as well as for firms.

- Why do insurance companies exist?

Example:
$$\text{Pr}(fire) = 1/250$$
Simulation

1. Simulate the damage of $n$ homeowners
2. Calculate the share
3. Repeat 1,000 times
4. Generate histogram

## Insurance Market
```{r,echo=FALSE}
ndraws    = 1000;
fireinsurance  = function(ninsur){
     fire = matrix(0,ndraws,1)
          for (i in 1:ndraws){
               fg      <- sample(x=c(1,0),size=ninsur,replace=TRUE,prob=c(1/250,249/250))
               fg      <- sum(fg)/ninsur
               fire[i] <- fg}
               return(fire)}
fire1     = fireinsurance(1000)
fire2     = fireinsurance(10000)
fire3     = fireinsurance(25000)
fire4     = fireinsurance(100000)
par(mfrow=c(2,2))
hist(fire1,xlim=c(0,0.025),ylim=c(0,750),main="1000 People Insured",xlab="",
          breaks=seq(0,0.03,0.001))
     hist(fire2,xlim=c(0,0.025),ylim=c(0,750),main="10000 People Insured",xlab="",
          breaks=seq(0,0.03,0.001))
     hist(fire3,xlim=c(0,0.025),ylim=c(0,750),main="25000 People Insured",xlab="",
          breaks=seq(0,0.03,0.001))
     hist(fire4,xlim=c(0,0.025),ylim=c(0,750),main="100000 People Insured",xlab="",
          breaks=seq(0,0.03,0.001))
```



