---
title: "Confidence Intervals"
author: "Jerome Dumortier"
date: "`r format(Sys.time(),'%d %B %Y')`"
output: 
     beamer_presentation:
          theme: "Hannover"
classoption: "aspectratio=1610"
linkcolor: MidnightBlue
---

```{r,echo=FALSE,warning=FALSE,results=FALSE}
load("G:/My Drive/Teaching/Data Analysis for Public Affairs/GitHub/DataAnalysisPAData.RData")
```

## Lecture Overview
Topics covered:

1. Definition of a confidence interval
2. Confidence interval for the mean
    - Known variance (unrealistic)
    - Unknown variance
3. Confidence interval for a proportion
4. Sample size calculation
    - Infinite population
    - Finite population

## Definition
Confidence interval:

- A range of values based on a sample such that the population parameter, i.e., mean or proportion, is occurring in the range with probability $\alpha$. The specific probability is called level of confidence and is in most cases set to 95%.
- Confidence interval = point estimate $\pm$ margin of error

Put differently, if you take 100 samples and construct a confidence interval as outlined in the next slides; in 95 cases, that confidence interval will contain the population mean or proportion. The confidence interval is influenced by

- Sample size $n$
- Population standard deviation $\sigma$ (often estimated by $s$)
- Level of confidence

## Confidence Interval for the Mean
Components for any confidence interval involving the mean:

- Sample size
- Sample mean

Components depending on what we know about the population standard deviation:

- Known population standard deviation
    - Population standard deviation: $\sigma$
    - $z$-value for a given confidence level
- Unknown population standard deviation
    - Estimate of the population standard deviation: $s$
    - *t*-value for a given confidence level

In what follows, only a confidence interval for the mean with *unknown* variance is considered.

## Confidence Interval for the Mean: Overview
Given that the standard deviation of the population is unknown, the confidence interval is constructed as follows:
$$\bar{x} \pm t_{\alpha,df} \cdot \frac{s}{\sqrt{n}}$$
Notes:

- Requires estimation of the population variance $s^2$ and standard deviation through $s$.
- Use of the *t*-distribution instead of the standard normal distribution:
    - $\alpha$ is the confidence level, e.g., 95%
    - $df$ are the degrees of freedom

## Confidence Interval for the Mean: Steps
Steps to construct the confidence interval if $\sigma$ is unknown:

1. Estimate the sample mean $\bar{x}$
2. Determine the degrees of freedom $df=n-1$
3. Determine $t_{\alpha,df}$: For a 95% confidence interval, this can be done with any statistical software:
    - In R: `qt(0.975,df=N-1)`
4. Use the equation:
    $$\bar{x} \pm t_{\alpha,df} \cdot \frac{s}{\sqrt{n}}$$

## Confidence Interval for the Mean: Example
Assume the American Economic Association (AEA) wants to construct a 95% confidence interval for the starting salaries of economics majors. The sample mean of 36 randomly selected graduates is \$48,500. The calculated sample variance is $s=\$3,600$.
$$\$48,500 \pm 2.03 \cdot \frac{\$3,600}{\sqrt{36}} = \$48,500 \pm \$1,218$$

The value of 2.03 leaves 2.5% in each tail of the *t*-distribution with 35 degrees of freedom. The \$1,218 represents the margin of error. The value of \$600 (i.e., $\$3,600=/\sqrt{36}$) represents the standard error.

## Confidence Interval for the Mean: R
Confidence Interval for the Mean

```{r,results=FALSE}
nobs = nrow(mh2)
meandata = mean(mh2$price)
stdev = sd(mh2$price)
t_alpha_df = qt(0.975,nobs-1)
CI_lower = meandata-t_alpha_df*stdev/sqrt(nobs)
CI_upper = meandata+t_alpha_df*stdev/sqrt(nobs)
t.test(mh2$price)
```

## Confidence Interval for a Proportion: Overview
For a proportion, we have the following:

- Estimate proportion from the data: $\hat{p}$
- Estimate standard deviation: $\sigma=\sqrt{\hat{p} \cdot (1-\hat{p})}$

So the standard error for the proportion is
$$\sigma_{\hat{p}}=\sqrt{\frac{\hat{p} \cdot (1-\hat{p})}{n}}$$

Thus, the 95% confidence interval is constructed as follows:
$$\hat{p} \pm 1.96 \cdot \sigma_{\hat{p}}  \Leftrightarrow \hat{p} \pm 1.96 \cdot \sqrt{\frac{\hat{p} \cdot (1-\hat{p})}{n}}$$

## Confidence Interval for a Proportion: Example
Assume that you are interested in the political party affiliation of voters. Suppose that $n=1000$ and $p_{GOP}= 0.55$. Given the equation on the previous slide, you can calculate: 
$$\sigma_{\hat{p}}=\sqrt{\frac{0.55 \cdot (1-0.55)}{1000}} = 0.0157$$
This gives us the margin of error for the political party affiliation of $0.55 \pm 1.96 \cdot 0.0157$.

## Confidence Interval for a Proportion: R
```{r,results=FALSE}
nobs = nrow(gss2018)
meandata = mean(gss2018$vote)
z = qnorm(0.975)
stderror = sqrt(meandata*(1-meandata)/nobs)
CI_lower = meandata-z*stderror
CI_upper = meandata+z*stderror
t.test(gss2018$vote)
```

## Sample Size Calculations
Recall from the confidence interval that
$$p = z \cdot \frac{\sigma}{\sqrt{n}}$$

Margin of error depends on sample size $n$. Possibility of calculation the sample size necessary to achieve a given margin of error. Two possible cases:

- Infinite population
- Finite population

Equation:
$$z \cdot \sqrt{\frac{\sigma^2}{n}} \leq \epsilon \quad \Rightarrow n \geq \left( \frac{z \cdot \sigma}{\epsilon} \right)^2$$

## Sample Size Calculations: Infinite Population
With prior knowledge about the proportion
$$n \geq \frac{z^2 \cdot p \cdot (1-p)}{\epsilon^2}$$

Without prior knowledge about the proportion
$$n \geq \left( \frac{z \cdot 0.5}{\epsilon} \right)^2$$

## Sample Size Calculations: Example
Suppose you want to know how many people support a property tax reform. You do not have any knowledge about the population parameters but want the estimate to be within 2%. For this reason, you adopt an initial estimate of $p=0.5$. This results in a "worst case" scenario.
$$ n= \left( \frac{1.96 \cdot 0.5}{0.02} \right)^2 =2401$$

## Sample Size Calculations: Finite Population
The sample size necessary also depends on the population size. Suppose you are interested in how many students support a privatization of parking.
$$n_f = \frac{n_{\infty} \cdot N}{n_{\infty} + (N-1)}$$
For a college with 10,000 students:
$$\frac{2401 \cdot 10000}{2401 +( 10000-1)}=1937$$
The sample size needed for large-sample confidence intervals is $n \cdot \hat{p} \geq 15$. Sometimes it is easier to go with a different rule thumb that specifies that $n \geq 30$.