---
title: "Hypothesis Testing"
author: "Jerome Dumortier"
date: "`r format(Sys.time(),'%d %B %Y')`"
output: 
     beamer_presentation:
          theme: "Hannover"
classoption: "aspectratio=169"
linkcolor: MidnightBlue
---

```{r,echo=FALSE,warning=FALSE,results=FALSE}
load("G:/My Drive/Teaching/Data Analysis for Public Affairs/GitHub/DataAnalysisPAData.RData")
```

## Example of Hypothesis Testing
Weight and weight distribution are important parameters for airplanes and the European Aviation Safety Agency (EASA) publishes passenger standard weights:

- Female passenger: 66.5 kilograms (146.6 lbs)
- Male passenger: 84.6 kilograms (186.5 lbs)
- Children under age 12: 30.7 kilograms (67.7 lbs)

In 2017, the airline Finnair asks passengers to step on a scale before boarding flights to better understand their payload. At the time of the BBC article [Why Finnair wants to put passengers on the weighing scales](https://www.bbc.com/news/business-41835933), they had collected data from 180 volunteers.

## Introduction to Hypothesis Testing
Hypothesis:

- A statement about a parameter taking on a particular value. This is formulated as the null hypothesis $H_0$.

Hypothesis test:

- A procedure to verify the hypothesis based on a random sample of size N. We never accept $H_0$ but **fail to reject** $H_0$. This is similar to guilty versus not guilty in courts. The opposite of the null hypothesis is labeled $H_a$ (sometimes $H_1$) as the alternative hypothesis.

## Overview of Hypothesis Tests
One-sample (or one-group) tests:

- Population mean with unknown variance
- Population proportion

Two-sample (or two-group) tests:

- Population proportions
- Population means
     - Equal versus unequal variance
- Paired difference test

Note:

- Statistics textbooks often include "population mean with known variance." This is a highly unlikely case and thus, it is skipped for this section.

## Hypothesis Testing Procedure
Steps

1. Formulating the null hypothesis $H_0$ stating that the parameter takes a particular value:
     - One-sided test: $H_0$: $\mu \geq \mu_0$ or $\mu \leq \mu_0$
     - Two-sided test: $H_0$: $\mu = \mu_0$

2. Setting the significance level $\alpha$, e.g., 1%, 5%, or 10%.
3. Test statistic: Value based on the sample used to **reject** or **fail to reject** the null hypothesis.
4. Critical value and *p*-value:
     - Critical value represents the border point between rejecting and failing to reject $H_0$.
     - *p*-Value: Probability of observing the parameter given the null hypothesis. Small *p*-values represent evidence against $H_0$.
     
Note that equality is always part of $H_0$, i.e., $=$, $\leq$, or $\geq$.

## Decisions and Errors in Hypothesis Testing

| Null Hypothesis | Fail to reject $H_0$ | Reject $H_0$ |
|:----------------|:--------------------:|:------------:|
| $H_0$ is true   | Correct              | Type I Error |
| $H_0$ is false  | Type II Error        | Correct      |

Type I Error:

- Probability of rejecting $H_0$ when it is true.
- Also known as the significance level of a test denoted with $\alpha$.

Type II Error:

- Probability of failing to reject $H_0$ when it is false.

## Interpretation of the *p*-Value
Each statistical software provides a *p*-value:

- Lowest level of significance at which the null hypothesis can be rejected.
- Represents the probability of observing the sample given that the hypothesis is true. The lower the *p*-value the more unlikely is the hypothesis.
- The null hypothesis $H_0$ is rejected if the *p*-value is smaller than the significance level.

The smaller the *p*-value, the stronger the evidence against $H_0$ being true. This is true for any type of hypothesis test.

## Mean with Unknown Variance: Test Statistic
Unknown variance requires the use of the *t*-distribution given the following test statistic:
$$t = \frac{\bar{x}-\mu_0}{s/\sqrt{n}}$$

where

- $\bar{x}$ is the sample mean
- $\mu_0$ is the hypothesized mean
- $s$ is the sample standard deviation
- $n$ is the sample size

## Two-Sided Test for Water Pressure
You work as a engineer for the local water company and you are concerned about the daily water pressure in the city's pipes. Too much pressure may burst pipes whereas too little pressure causes customer complaints. Regulation requires a water pressure of 50 psi. You collect a sample of 30 daily water pressures. The sample mean and the sample standard deviation are $\bar{x}=51.788$ and $s=3.389$.

## Two-Sided Test for Water Pressure: Step-by-Step
Formulating the null hypothesis $H_0$ and the alternative hypothesis $H_a$

- $H_0$: $\mu = 50$
- $H_a$: $\mu \neq 50$

Setting the significance level:

- $\alpha = 0.05$ or 5\%
   
Calculating the test statistic
$$t = \frac{51.788-50}{3.389/\sqrt{30}} = 2.8895$$

The critical value for $t_{0.05/2,29}$ is $2.045$ (`qt(c(0.025,0.975),df=29)`). Thus, we reject $H_0$ and there is evidence that the water pressure is different from 50 psi. To calculate the $p$-value in R: `(1-pt(tstatistic,29))*2`

## Two-Sided Test for Water Pressure: R
Note that the data in `waterpressure` is randomly generated and value differ from previous slide.
```{r}
t.test(waterpressure$psi,mu=50)
```

## One-Sided Test for MPA Scores
Consider the scores from a graduate MPA class which has eighteen students in `mpa`.

- Sample mean: $\bar{x} = 69$
- Sample standard deviation: $s=21.14933$.
            
We are interested in the null hypothesis $H_0$: $\mu \geq 80$. We can compute the *t*-statistic as follows:
$$t = \frac{69-80}{21.14933/ \sqrt{18}} = -2.206644$$
What is the critical value in this case?

## One-Sided Test for MPA Scores
```{r}
t.test(mpa$scores,mu=80,alternative="less")
```

## Hypothesis Test about Population Proportion
Test statistic for a proportion
$$z=\frac{\bar{p}-p_0}{\sqrt{p_0 \cdot (1-p_0)/n}}$$
where $p_0$ is the hypothesized population proportion. Recall that the sampling distribution of a sample proportion has mean $p$ and standard error $\sqrt{p \cdot (1-p)/n}$. Example:

- Use of Instagram by GSS respondents
- $\bar{p}=0.3097$ and $n=1366$

Let us calculate two possible hypothesis tests manually and with R.

## Hypothesis Test about Population Proportion
Example: Assume $\bar{p}=0.3097$ and $n=1366$. Under $H_0$: $p_0=0.33$, the test statistic is
$$z=\frac{0.3097-0.333}{\sqrt{\frac{0.333(1-0.333)}{1366}}}=-1.8914$$
For a two-sided hypothesis test at the $\alpha=0.05$ level, we fail to reject the hypothesis because $1.8914<1.96$.

## Hypothesis Test about Population Proportion: Social Media
```{r}
gss$instagram    = NA
gss$instagram[which(gss$instagrm=="no")]     = 0
gss$instagram[which(gss$instagrm=="yes")]    = 1
t.test(gss$instagram,mu=0.33,na.action=na.omit)
```

## Hypothesis Test about Population Proportion: Social Media
```{r}
t.test(gss$instagram,mu=0.33,
       alternative="less",na.action=na.omit)
```

## Two Sample Test: Overview
Difference between two mean:
$$\bar{x}_1 - \bar{x}_2$$

Means of two dependent populations:

- Assumption of equal variance, i.e., $\sigma^2_1 = \sigma^2_2$
- Example: Pre- and post-test
- Pooled-Variance t-test: One estimate of unknown $\sigma^2$, i.e., $s_p$

Means of two independent populations:

- Assumption of unequal variance, i.e., $\sigma^2_1 \neq \sigma^2_2$
- Samples from two different populations
- Separate-Variance t-test: Two estimates for unknown $\sigma^2_1$ and $\sigma^2_2$.

## Two Population Means: Equal Variance
Pooled variance:
$$s_p^2 = \frac{(n_1-1) \cdot s^2_1+(n_2-1) \cdot s^2_2}{n_1+n_2-2}$$
Test statistic:
$$t = \frac{(\bar{x}_1-\bar{x}_2)-(\mu_1-\mu_2)}{\sqrt{s_p^2 \cdot \left(\frac{1}{n_1} +\frac{1}{n_1} \right)}}$$
Confidence interval for $\mu_1-\mu_2$:
$$(\bar{x}_1-\bar{x}_2) \pm t_{\alpha/2} \sqrt{s_p^2 \left(\frac{1}{n_1} +\frac{1}{n_1} \right)}$$

## Two Population Means: Unequal Variance
Test statistic:
$$t = \frac{(\bar{x}_1-\bar{x}_2)-(\mu_1-\mu_2)}{\sqrt{\left(s_1^2 /n_1 +s_2^2/n_2 \right)}}$$
Degrees of Freedom:
$$d_1 = \left(s_1^2/n_1+s_2^2/n_2 \right)^2$$
$$d_2 = \left(s_1^2 /n_1\right)^2$$
$$d_3 = \left(s_2^2 /n_2\right)^2$$
Then the degrees of freedom are $d.f. = d_1/(d_2+d_3)$.

## Two Sample Test (Equal and Unequal Variance): R
Separating schools into three different income groups
```{r,results=FALSE}
oh   = merge(ohioincome,ohioscore,by="irn")
oh_s = subset(oh,enrollment<1000)
oh_l = subset(oh,enrollment>3000)
oh_m = subset(oh,enrollment>1000 & enrollment <3000)
t.test(oh_s$score,oh_l$score,var.equal = TRUE)
t.test(oh_s$score,oh_l$score,var.equal = FALSE)
```

## Paired Difference Test: Related Populations
Example: Textbook prices:

- Online vs. bookstore because prices exist for both purchase options.

Difference between paired (!) values:
$$D_i = x_{1,i} - x_{2,i}$$
Elimination of variation among subjects. Point estimate for paired difference
$$\bar{D} = \frac{1}{n} \sum_{i=1}^{n} D_i$$
Sample standard deviation
$$S_d = \sqrt{\frac{\sum_{i=1}^{n} (D_i-\bar{D})^2}{n-1}}$$

## Paired Difference Test: Related Populations
Test statistic:
$$t_p = \frac{\bar{D}-\mu_D}{S_d/\sqrt{n}}$$
Confidence interval:
$$\bar{D} \pm t_{\alpha/2} \frac{S_D}{\sqrt{n}}$$
$t_p$ has $n-1$ degrees of freedom

## Textbook Example

| Book        | Online | Bookstore | Difference |
|-------------|-------:|----------:|-----------:|
| History 1   |   10.2 |      11.4 |       -1.2 |
| History 2   |  18.95 |        19 |      -0.05 |
| Economics 1 | 184.53 |    200.75 |     -16.22 |
| Business 1  | 236.75 |     247.2 |     -10.45 |
| Business 2  |  67.41 |     71.25 |      -3.48 |

Note that $\sum D_i = -31.76$, $\bar{D} = -6.352$, and $s_D = 6.833$.

## Textbook Example
```{r,results=FALSE}
online      = c(10.20,18.95,184.53,236.75,67.41)
bookstore   = c(11.40,19,200.75,247.20,71.25)
t.test(online,bookstore,paired=TRUE)
```

## Hypothesis Tests for Population Proportions
Confidence interval and hypothesis test for difference between two population proportions. Point estimate for difference:
$$p_1-p_2$$
Pooled estimate for overall proportion
$$\bar{p} = \frac{x_1+x_2}{n_1+n_2}$$
Test statistic
$$z_p = \frac{(p_1-p_2)-(\eta_1-\eta_2)}{\bar{p}(1-\bar{p})\left( \frac{1}{n_1} + \frac{1}{n_2} \right)}$$

## Hypothesis Tests for Population Proportions
Gun ownership among males and females from the General Social Survey

- Females: 495 no, 207 yes, $n_1 = 702$
- Males: 334 no, 231 yes, $n_2 = 565$

Is there a statically significant difference in gun ownership between women and men?
$$\bar{p} = (207+231)/(702+565) = 0.3457$$
$$p_1 = 207/702 = 0.2949$$
$$p_2 = 231/565 = 0.4088$$

