---
title: "Introduction to Statistics"
author: "Jerome Dumortier"
date: ""
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Uncertainty, Probability, and Statistics in Everyday Life
Grades

- Uncertainty surrounding class grade during a semester
- Association of probabilities with each grade

Fire station calls

- Number and location of calls
- Number of fire trucks and other vehicles required

Two outcomes does not mean a 50\% chance for each to happen

- Success of a free throw by Stephen Curry
- Flight delay due to fog

Netflix

- Recording of all activities, e.g., browsing, pausing, rewinding
- 75\% of viewings are based on recommendations by Netflix
- [Do You Know When You Were Hooked? Netflix Does](https://media.netflix.com/en/press-releases/do-you-know-when-you-were-hooked-netflix-does)

## Some Examples of Statistics in the News I
Election outcomes

   - 2016 U.S. Presidential Election
     - FiveThirtyEight 2016 Election Forecast (chance of winning): Donald J. Trump 28.6\%
     - [One explanation for the surprising result](http://news.harvard.edu/gazette/story/2017/03/nate-silver-says-conventional-wisdom-not-data-killed-2016-election-forecasts/)
   - 2002 French Presidential Election
     - Two-stage election
     - Final round: Jacques Chirac (82.2\%) and Jean-Marie Le Pen (17.8\%)

## Some Examples of Statistics in the News II
Evolution of the stock market

- Importance of correlation among stocks and mutual funds

Path of a hurricane:
        \begin{itemize}
            \item \href{https://www.weather.gov/okx/HurricaneSandy}{Sandy (National Weather Service NWS)}
            \item \href{https://www.nhc.noaa.gov/archive/2012/graphics/al18/loop_S.shtml}{Sandy (National Oceanic and Atmospheric Administration NOAA)}
        \end{itemize}

## Slide with Plot

```{r pressure}
plot(pressure)
```

