---
title: "Introduction to R/RStudio"
author: "Jerome Dumortier"
date: "`r format(Sys.time(),'%d %B %Y')`"
output: 
     beamer_presentation:
          theme: "Hannover"
          slide_level: 3
section-titles: false
classoption: "aspectratio=169"
linkcolor: MidnightBlue
---

## Tutorial Overview
```{r DataLoad,echo=FALSE}
load("G:/My Drive/Teaching/Data Analysis for Public Affairs/GitHub/DataAnalysisPAData.RData")
```

Topics covered in this tutorial:
        
- Overview of R/RStudio
- Data management
- Plotting and graphs with R
- Basic statistics
        
Exercises will be conducted throughout the tutorial.

# Overview of R/RStudio
## Online Resources and Help
Very large user community for R

- Google search for "Some topic R" usually leads quickly to the desired help.

Here are the links to a few online tutorials:

- [UCLA OARC Statistical Methods and Data Analytics](https://stats.oarc.ucla.edu/r/)
- [Statmethods](www.statmethods.net)

There is also `www.sthda.com/english/`, which is very useful for some more advanced applications like plotting with the package `ggplot2`.

## Two Particularly Useful Online Resources
Two online resources will provide you the solution to the vast majority of your R questions. Getting to those websites is usually the result of a Google search.

- [Statistical Data Analysis R](https://stat.ethz.ch/R-manual/): This resource contains the function manual for R/RStudio including all packages:

     - Example for [boxplot](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/boxplot.html)
     - The most helpful part are the examples at the bottom of the page.

- Stack Overflow: Resources for developers
     
     - Google search: [r ggplot two y axis](https://stackoverflow.com/questions/3099219/plot-with-2-y-axes-one-y-axis-on-the-left-and-another-y-axis-on-the-right)

Note that all questions on Stack Overflow have to be accompanied by an easily reproducible example.

## Opening RStudio
Working with RStudio is done in four windows:
        
- Script Window
 
     - This is were you type your R Script (.R) and where you execute commands.
     - Comparable to do-file/editor in Stata.
     - This window needs to be opened by File $\Rightarrow$ New File $\Rightarrow$ R Script.
 
- Console window
 
     - Use of R interactively. Should only be used for quick calculations and not part of an analysis.
 
- Environment
 
     - Lists all the variables, data frames, and user-created functions.
     - It is tempting to use the "Import Dataset" function \dots Don't!
 
- Plots/Packages/Help

## Packages
There is a base version of R that allows doing many calculations but the power of R comes through its many packages. To use functions associated with a particular package (e.g., to read data from Excel), click "Install" in the packages window of RStudio and type in the name of the desired package. Or alternatively, use

`install.packages("openxlsx")`

To use a package, you have to activate it by including:

`library("openxlsx")`

Packages are updated on a regular basis by users.

## Front Matter
The purpose of the hashtag is that R will skip whatever is after. The following command clears all variables from R:

`rm(list=ls())`

To display the current working directory and to set a new one:

`getwd()`

`setwd("C:/Users/Jerome/Documents/R Lecture")`

You have to change the part between the quotation marks to the directory you have created. For file paths, replace backslashes with forward slashes. The following  will import sample data. 

`honda = read.csv("honda.csv")`

It is also good practice to save your R-script on a regular basis.

## Exercise 1
Create a R-script file with the following components:
        
- Two lines for the title and the date (use \#)
- Clearing all current contents
- Setting the correct working directory. This should be a folder to which you have downloaded all materials.
- Installing and loading the package \textcolor{blue}{openxlsx}.

## Functions I
At the core of R are functions that "do things" based on your input. The basic structure is

`object = functionname(argument1=value,argument2=value,...)`

Components
    
- `object`: Output of the function will be assigned to object.
- `functionname`: Name of the system function. You can also create and use your own functions. More about this later.
- `argument`: Arguments are function specific.
- `value`: The value you want a particular argument to take.
    
## Functions II
Notes:
        
- If a function is executed without an specific assignment, the output will be displayed in the console window.
- Before using a function, read the documentation.
- Many functions have default settings. Be aware of default values. In most cases, those defaults are set to values that satisfy most uses.
        
Notation in the help file:
        
- Consider the help file for the function [hist](https://stat.ethz.ch/R-manual/R-devel/library/graphics/html/hist.html)
        
Example about default values:
        
- \texttt{t.test(x,\textcolor{red}{y=NULL},[\dots],\textcolor{red}{mu=0},\textcolor{red}{conf.level=0.95},[\dots])}
    
# Data Management    
## Data in R
The main data types in R are:
        
- Vectors
 
     - `preselection = seq(1788,2016,4)`
     - `midterm = seq(by=4,to=2018,from=1790)`
 
- Matrix (only numerical values are allowed)
 
     - `somematrix = matrix(8,10,4)`

- Data frames
 
     - By far, the most common data type in R.
     - Comparable to an Excel sheet.
     - More on this later.
 
- Lists (Collection of objects from of various types)
 
     - `myfirstlist = list(preselection,midterm,somematrix)`
                
## Using R as a Calculator
Entering heights of people and storing it in a vector named `height`:
```{r}
height = c(71,77,70,73,66,69,73,73,75,76)
```

Calculating the sum, product, natural log, or mean is done with the following commands:
```{r,results='hide'}
sum(height)
prod(height)
log(height) # Default is the natural log
meanheight = mean(height)
```

Calculating the height squared (element-wise squaring):
```{r,results='hide'}
height_sq = height^2
```

Removing (i.e., deleting) unused elements: `rm(heightsq,meanheight)`

## Creating a Data Frame from Scratch
Data frames are the most commonly used tables in R/RStudio. They are similar to an Excel sheet.
    
- Column names represent the variables and rows represent observations.
- Column names must be unique and without spaces.
    
Suggestion: Use only lower-case variable names and objects.
```{r,results='hide'}
studentid      = 1:10
studentnames   = c("Andrew","Linda","William","Daniel",
                   "Gina","Mick","Sonny","Wilbur",
                   "Elisabeth","James")
students       = data.frame(studentid,studentnames,height)
rm(studentid,height,studentnames)
```

## Exercises 2
Create a data frame called `students` containing the following information:

| Name    | Economics | English |
|---------|----------:|--------:|
| Mindy   |      80.0 |    52.5 |
| Ruiqing |      60.0 |    60.0 |
| Shubra  |      95.0 |    77.5 |
| Keith   |      77.5 |    30.0 |
| Luisa   |      97.5 |    95.0 |
        
- Use *name* as the column header for the students' names.
- Once you have created the data frame, remove the unused vectors.
        
## Indexing
Indexing refers to identifying elements in your data:
        
- For most objects: \texttt{students[row number,coloumn number]}
                
     - `students[3,2]` returns 95
     - What does `students[3,]` return?
 
- If you want to select certain columns: `students[c("name")]`
 
     - Other example: `students[c("name","english")]`
 
- Selecting results based on certain conditions: `students[which(students$economics>80),]`

Referring to a particular column in a data frame is done through the dollar symbol:

- `students$english`
- You will use this functionality very often.

Creating a new column: `students$average = rowMeans(students[c("economics","english")])`

## Basic Handling of Data Frames I
Data on vehicle fuel efficiency for all model years (1984--2020) from [DOE and EPA](https://www.fueleconomy.gov/feg/download.shtml) with corresponding [documentation](https://www.fueleconomy.gov/feg/ws/index.shtml#vehicle) of the variables. Sub-setting data is done with the command `subset`:

```{r,results='hide'}
cars2015 = subset(vehicles,year==2015)
```

Note that the double equal sign conducts a logical test. To list all EPA vehicle size classes ($vclass$):

```{r,results='hide'}
unique(cars2015$vclass)
```

Suppose you are only interested in the variables *ghgScore* and *VClass* for the model year 2015.
        
- `cars2015 = subset(vehicles,year==2015,select=c("ghgScore","VClass"))`
        
Get glimpse at the results: `table(cars2015$vclass,cars2015$ghgScore)`

## Basic Handling of Data Frames II
Suppose you are only interested in *Compact Cars* and *Large Cars* in the column *VClass* for the year 2015. There the notation is a bit odd (note that the many line breaks are not necessary to include in R):

```{r,results='hide'}
cars2015 = subset(vehicles,
                  year==2015 &
                  vclass %in% c("Compact Cars",
                  "Large Cars"),
                  select=c("make","co2tailpipegpm","vclass"))
```

## Exercises 3
From the vehicles data set, extract the GHG Score and the vehicle class from the 2014 model year for the following manufacturers: Toyota, Ford, and Audi. Your new data set should contain the following columns: *ghgScore*, *make*, and *VClass*. Is the resulting data frame sensible or do you see a problem?

## Aggregating Data and Writing .csv-Files
To aggregate data based on a function, e.g., sum or mean:
```{r}
cars2015 = aggregate(cars2015$co2tailpipegpm,
                     by=list(cars2015$make,cars2015$vclass),
                     FUN=mean)
```

To write data to the current working directory:

`write.csv(cars2014,"cars2014.csv")`

## Importing Data into R/RStudio
Machine-readable data can be imported as follows:
        
- \texttt{read.csv("filename.csv")}: If you have a comma separated value (.csv) file then this is the easiest and preferred way to import data.
- \texttt{readWorkbook(file="filename.xlsx",sheet="sheet name")}: Requires the package \textcolor{blue}{openxlsx}. Note that there are many packages reading Excel and this is the most reliable and user-friendly.
        
Importing data from other software packages (e.g., SAS, Stata, Minitab, SPSS) or .dbf (database) files:
        
- Package \textcolor{blue}{foreign} reads .dta Stata files (Version 5-12) with the command \texttt{read.dta}
- Package \textcolor{blue}{readstata13} reads files from newer Stata versions
        
## Extending the Basic `table()` Function
Required package:
        
- \textcolor{blue}{gmodels}
        
Compare the outputs of `table()` and `CrossTable()`

```{r,results='hide'}
library(gmodels)
table(gss$owngun,gss$sex)
CrossTable(gss$owngun,gss$sex)
```

Note that for almost any R command, you can store the output by assigning it a name:
```{r,results='hide'}
output = CrossTable(gss$owngun,gss$sex)
```

## Merging Datasets (`ohioscore` and `ohioincome`)
Consider two datasets from school districts in Ohio:
        
- \texttt{ohioscore} which contains an identifier column \emph{IRN} and a score that indicates quality of the school.
- \texttt{ohioincome} which contains the same identifier than the previous sheet in addition to median household income and enrollment.
        
One important function to merge datasets in R:
```{r}
ohioschool = merge(ohioscore,ohioincome,by=c("irn"))
rm(ohioscore,ohioincome)
```

# Plotting and Graphs with R
## Overview
Before we start talking about graphics, execute the following command: 

`demo(graphics)`

R has very advanced graphing capabilities that allows you to do any type of visualization. Personally, I use it most often for:
    
- Automatically updating graphs for manuscripts
- Side-by-side plots
- Plotting maps
    
In almost all cases, vector graphs are preferred over bitmap graphs.

## Faithful Dataset: Summary
R and some packages include example data sets to facilitate learning the package. A "famous" R data set is `faithful`:
```{r}
faithful = faithful
summary(faithful)
cor(faithful$eruptions,faithful$waiting)
```

## Faithful Dataset: Histogram Setup
```{r,fig.show='hide'}
par(mfrow=c(1,2))
hist(faithful$eruptions,main="Eruption Time",
    xlab="Minutes",xlim=c(1,6),ylim=c(0,80))
hist(faithful$waiting,main="Waiting Time",
    xlab="Minutes",xlim=c(40,100),ylim=c(0,60))
```

## Faithful Dataset: Histogram Plot
```{r,echo=FALSE,fig.align='center',fig.height=3,fig.width=5}
par(mfrow=c(1,2))
hist(faithful$eruptions,main="Eruption Time",
     xlab="Minutes",xlim=c(1,6),ylim=c(0,80))
hist(faithful$waiting,main="Waiting Time",
     xlab="Minutes",xlim=c(40,100),ylim=c(0,60))
```

## Faithful Dataset: Correlation
```{r,echo=FALSE,fig.align='center',fig.height=3,fig.width=5}
plot(faithful$eruptions,faithful$waiting)
```

## Faithful Dataset: ECDF Setup
```{r,fig.show='hide'}
par(mfrow=c(1,2))
plot(ecdf(faithful$eruptions),
   main="Eruption Time",
   xlab="Minutes",xlim=c(1,6))
plot(ecdf(faithful$waiting),
   main="Waiting Time",
   xlab="Minutes",xlim=c(40,100))
```

## Faithful Dataset: ECDF Plot
```{r,echo=FALSE,fig.align='center',fig.height=3,fig.width=5}
par(mfrow=c(1,2))
plot(ecdf(faithful$eruptions),
   main="Eruption Time",
   xlab="Minutes",xlim=c(1,6))
plot(ecdf(faithful$waiting),
   main="Waiting Time",
   xlab="Minutes",xlim=c(40,100))
```

## Plotting Ohio School Scores: Setup
```{r plotohschools1,results='hide',message=FALSE,warning=FALSE,fig.show='hide'}
library(ggpubr)
library(ggsci)
topandbottom   = quantile(ohioschool$medianincome,
                          seq(0,1,0.25))
quartiles      = as.integer(cut(ohioschool$medianincome,
                            quantile(ohioschool$medianincome,
                            probs=0:4/4),include.lowest=TRUE))
ohioschool$quartiles                       = quartiles
ohioschool$income                          = NA
ohioschool$income[ohioschool$quartiles==1] = "Lower"
ohioschool$income[ohioschool$quartiles==2] = "Lower Mid."
ohioschool$income[ohioschool$quartiles==3] = "Upper Mid."
ohioschool$income[ohioschool$quartiles==4] = "Upper"
ggdensity(ohioschool,x="score",add="mean",
    color="income",fill="income",palette="jco")
```

## Plotting Ohio School Scores: Figure
```{r plotohschools2,echo=FALSE,fig.align='center',fig.height=3,fig.width=5}
ggdensity(ohioschool,x="score",add="mean",color="income",fill="income",palette="jco")
```