---
title: "Limited Dependent Variable Models"
author: "Jerome Dumortier"
date: "`r format(Sys.time(),'%d %B %Y')`"
output: 
     beamer_presentation:
          theme: "Hannover"
classoption: "aspectratio=169"
linkcolor: MidnightBlue
---

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
load("G:/My Drive/Teaching/Data Analysis for Public Affairs/GitHub/DataAnalysisPAData.RData")
library(AER)
library(censReg)
library(foreign)
library(MASS)
library(pscl)
library(stargazer)
library(survival)
library(survminer)
library(truncreg)
library(vcdExtra)
```

# Overview
## Packages and Files
Required packages:
        
- [AER](https://cran.r-project.org/web/packages/AER/index.html)
- [censReg](https://cran.r-project.org/web/packages/censReg/index.html)
- [foreign](https://cran.r-project.org/web/packages/foreign/index.html)
- [MASS](https://cran.r-project.org/web/packages/MASS/index.html)
- [pscl](https://cran.r-project.org/web/packages/pscl/index.html)
- [stargazer](https://cran.r-project.org/web/packages/stargazer/index.html)
- [survival](https://cran.r-project.org/web/packages/survival/index.html)
- [survminer](https://cran.r-project.org/web/packages/survminer/index.html)
- [truncreg](https://cran.r-project.org/web/packages/truncreg/index.html)

Required files:
```{r}
data("NMES1988",package="AER")
```

## Topics Covered
Regression models in which the dependent variable is somehow limited:

- Truncated data: Values above and/or below particular points are not reported
- Censored data: Values above and/or below particular points are reported at those points
- Count data: Discrete, integer count value 
- Survival/duration data: Time to a certain event 

# Truncation
## Overview
Concept

- Value above and/or below a certain point are not part of the data 

Examples

- Low income household studies
- On-site visitation data (unobserved non-visitors)
- Employment data on hours worked (excludes unemployed)

Simulated data

- "True" Coefficients: $\beta_0=-2$ and $\beta_1=0.5$
- Values $y<0$ are not reported in the data

Next slide: The green regression line is "correct" whereas the "red" is the line obtained from a regression model which ignores the truncation.  

## Graphical Illustration 
```{r,echo=FALSE,fig.width=10,fig.height=5}
bhat_real           = lm(y_real~x,data=truncation)
bhat_truncated      = lm(y_obs~x,data=truncation)
plot(truncation$x,truncation$y_obs,xlim=c(0,10),ylim=c(-3,5),pch=19,ylab="y",xlab="x")
     points(truncation$x,truncation$y_real)
     abline(bhat_real,col=c("darkgreen"))
     abline(bhat_truncated,col=c("red"))
```

## Setup for `truncation` Data
```{r}
truncation1    = truncation[c("y_real","x")]
truncation2    = subset(truncation,y_obs>0,select=c("y_obs","x"))
bhat_real      = lm(y_real~x,data=truncation1)
bhat_truncated = lm(y_obs~x,data=truncation2)
```

Required package to estimate a truncated model

-  [truncreg](https://cran.r-project.org/web/packages/truncreg/index.html)

Additional variable output $sigma$:

- Related to the truncated normal distribution

## Complete vs. Truncated Data with Regular OLS
\tiny
```{r,echo=FALSE}
stargazer(bhat_real,bhat_truncated,type="text",no.space=TRUE)
```
\normalsize 

## Results: Correcting for Truncation
\tiny
```{r,echo=FALSE}
bhat_correct = truncreg(y_obs~x,data=truncation2,
                        point=0,direction="left")
summary(bhat_correct)
```
\normalsize

## Achievement Scores: Data Load and Description
Loading the data using the package [foreign](https://cran.r-project.org/web/packages/foreign/index.html)
```{r}
url         = "https://stats.idre.ucla.edu/stat/data/truncreg.dta"
achievement = read.dta(url)
```

Description of the data from [UCLA Source](https://stats.oarc.ucla.edu/r/dae/truncated-regression/):

*"A study of students in a special GATE (gifted and talented education) program wishes to model achievement as a function of language skills and the type of program in which the student is currently enrolled. A major concern is that students are required to have a minimum achievement score of 40 to enter the special program. Thus, the sample is truncated at an achievement score of 40."*

## Achievement Scores: Regular OLS Estimation
\tiny
```{r,echo=FALSE}
bhat_ols = lm(achiv~langscore+prog,data=achievement)
summary(bhat_ols)
```
\normalsize

## Achievement Scores: Truncated Model
\tiny
```{r,echo=FALSE}
bhat_tr = truncreg(achiv~langscore+prog,data=achievement,
              point=40,direction="left")
summary(bhat_tr)
```
\normalsize

# Censoring
## Overview
Concept

- Value above and/or below a certain point are not part of the data 

Examples

- Capacity constrained data, e.g., class enrollments or ticket sales
- Hours worked (or leisure demand), which is essentially capacity constrained
- Commodity purchases (non-negative)

Simulated data

- "True" Coefficients: $\beta_0=-2$ and $\beta_1=0.5$
- Values $y<0$ are reported at 0

R package [censReg](https://cran.r-project.org/web/packages/censReg/index.html) to reduce bias

## Graphical Illustration
```{r,echo=FALSE,fig.width=10,fig.height=5}
bhat_real = lm(yreal~x,data=censoring)
bhat_censored = lm(y~x,data=censoring)
plot(censoring$x,censoring$yreal)
     points(censoring$x,censoring$y,pch=19)
     abline(bhat_real,col=c("darkgreen"))
     abline(bhat_censored,col=c("red"))
```

## Complete vs. Censored Data with Regular OLS
\tiny
```{r,echo=FALSE}
stargazer(bhat_real,bhat_censored,type="text",no.space=TRUE)
```
\normalsize

## Estimation of a Censored Model
\tiny
```{r,echo=FALSE}
b_correct = censReg(y~x,data=censoring)
summary(b_correct)
```
\normalsize

# Count Models
## Overview
Dependent variable

- Discrete, integer count data

Examples

- What are the number of arrests for a person?
- What determines the number of credit cards a person owns? 

Three count data models

1. Poisson regression 
2. Quasi-Poisson Regression Model
2. Negative Binomial Regression Model

Choice criteria: Presence or absence of overdispersion

- Overdispersion Variance of the dependent variable is larger than its mean.
- Poisson model is not suitable for overdispersion

## Packages
The main package used is [pscl](https://cran.r-project.org/web/packages/pscl/index.html). There is also an additional resource with more theoretical details on the topic: [Regression Models for Count Data in R](http://dx.doi.org/10.18637/jss.v027.i08). A more up-to-date version of the document may be found with the [pscl](https://cran.r-project.org/web/packages/pscl/index.html) package documentation. 

## Poisson Regression Model
Recall Poisson distribution:
$$Pr(Y=k)=\frac{e^{-\lambda} \cdot \lambda^k}{k!}$$
Equidispersion as key characteristics:

- Mean and variance equal to $\lambda$, i.e., $E(Y)=\lambda$ and $Var(Y)=\lambda$
- Poisson regression: $\lambda=exp(\beta_0+\beta_1 \cdot x_1+ \dots + \beta_k \cdot x_k)$.

## NHTS Example: Number of Vehicles (`hhpub`)
Data source

- 2017 [National Household Travel Survey](https://nhts.ornl.gov/) 
- Survey quantifying trip and travel habits across the United States
- Example use: Quantifying intra-day electricity demand from electric vehicles

Outcome of interest

- Number of vehicles based on household income, home ownership, and urban/rural household location

Data preparation

- Elimination of missing and unknown data value
- Conversion of income to 1,000 dollars

## Data Preparation
```{r poissonregression_cleanhhpub,results=FALSE}
hhpubdata           = subset(hhpub,HHFAMINC %in% c(1:11) & 
                                   HOMEOWN %in% c(1,2) & 
                                   URBRUR %in% c(1,2) & 
                                   HHVEHCNT %in% c(0:12))
HHFAMINC            = c(1:11)
INCOME              = c(10,12.5,20,30,42.5,57.5,82.5,112.5,137.5,
                        175,200)
INCOME              = data.frame(HHFAMINC,INCOME)
hhpubdata           = merge(hhpubdata,INCOME)
hhpubdata$RURAL     = hhpubdata$URBRUR-1
hhpubdata$RENT      = hhpubdata$HOMEOWN-1
```

## Poisson Model Execution
Preliminary step: Calculation of mean and variance of dependent variable
```{r}
mean(hhpubdata$HHVEHCNT)
var(hhpubdata$HHVEHCNT)
```

Similar values and thus, Poisson regression model as an appropriate first step. 

```{r}
bhat_pois = glm(HHVEHCNT~INCOME+RENT+RURAL,
                data=hhpubdata,family=poisson)
```

## Results
\tiny
```{r,echo=FALSE}
summary(bhat_pois)
```
\normalsize

## Interpretation
Sign of coefficients as an indication of the direction of influence on the outcome variable, i.e., the number of cars. 

- Association of higher income and rural living with a higher number of car
- Association of renting with lower number of vehicles. 
- Possible correlation between income and renting 

General coefficient interpretation using $\exp(\beta)$, i.e., every unit increase in $X$ has a multiplicative effect of $\exp(\beta)$ on the mean of $Y$, i.e., $\lambda$:

- $\beta=0 \Rightarrow \exp(\beta)=1$: $Y$ and $X$ are not related.
- $\beta>0 \Rightarrow \exp(\beta)>1$: Expected count $E(y)$ is $\exp(\beta)$ times larger than when $X = 0$
- $\beta<0  \Rightarrow \exp(\beta)<1$: Expected count $E(y)$ is $\exp(\beta)$ times smaller than when $X = 0$

## Testing for Overdispersion I
Function `dispersiontest()` from the package [AER](https://cran.r-project.org/web/packages/AER/index.html):

- Tests the null hypothesis of equidispersion (i.e., assuming no overdispersion)

Executed after the main regression using `glm(...,family=poisson)` 

## Testing for Overdispersion II
```{r}
dispersiontest(bhat_pois)
```

Given the *p*-value, the null hypothesis cannot be rejected. If the data suggests overdispersion, two alternative regression models can be used: (1) Quasi-Poisson and (2) Negative Binomial.

## Quasi-Poisson Regression Model
Dataset `blm` from article [Black Lives Matter: Evidence that Police-Caused Deaths Predict Protest Activity](https://doi.org/10.1017/S1537592717004273). 

- Dependent variable: Total number of protests in a city
- Note that the paper includes a significant number of supplementary materials which allows for the replication of the results and much more.  

First step: Calculation of mean and variance of the variable $totalprotests$:

```{r}
mean(blm$totprotests)
var(blm$totprotests)
```

## Presence of Overdispersion
Likely overdispersion due to variance being significantly higher than mean. In a first step, a regular Poisson model is estimated.

```{r}
eq1       = "totprotests~log(pop)+log(popdensity)+percentblack+
             blackpovertyrate+I(blackpovertyrate^2)+
             percentbachelor+collegeenrollpc+demshare"
eq2       = paste(eq1,"+deathsblackpc",sep="")
eq3       = paste(eq1,"+deathspc",sep="")
bhat1     = glm(eq1,data=blm,family=poisson)
bhat2     = glm(eq1,data=blm,family=quasipoisson)
```

## Estimation Results
\tiny
```{r,echo=FALSE}
stargazer(bhat1,bhat2,type="text",no.space=TRUE,single.row=TRUE)
```
\normalsize
Note: Switch of statistical significance for population density 

## Testing for Overdispersion
```{r,echo=FALSE}
dispersiontest(bhat1)
```
Null hypothesis rejected at 10\% but not 5\% significance level. The Quasi-Poisson Regression Model handles overdispersion by adjusting standard errors but leaving the coefficient estimates the same.

## Negative Binomial Regression Model
The Negative Binomial Regression Model can be used in the presence of count data and overdispersion. Below, the results from the article [Black Lives Matter: Evidence that Police-Caused Deaths Predict Protest Activity](https://doi.org/10.1017/S1537592717004273) are recreated using the negative binomial models presented in the paper. 

Three models:

1. Resource mobilization and opportunity structure
2. Adding black death
3. Adding all police-caused deaths instead (victims of any race)

## BLM Models
```{r}
bhat3 = glm.nb(eq1,data=blm,link=log)
bhat4 = glm.nb(eq2,data=blm,link=log)
bhat5 = glm.nb(eq3,data=blm,link=log)
```

## BLM Model Results
\tiny
```{r,echo=FALSE}
stargazer(bhat3,bhat4,bhat5,type="text",no.space=TRUE,single.row=TRUE)
```
\normalsize

# Hurdle and Zero-Inflation Models
## Overview
Problem:

- Presence of many observations at 0 in count data
- Issues using Poisson or a Negative-Binomial Regression Model. 

Application of hurdle and zero-inflated models:

- Data `NMES1988` from the package [AER](https://cran.r-project.org/web/packages/AER/index.html) 
- Data BLM protests 

`NMES1988` Data:

- 4406 observations of people on Medicare who are 66 years or older. 
- Outcome of interest: Number of doctor $visits$ 
- Independent variables: $hospital$ (number of hospital visits), $health$ (self-indicated health status), $chronic$ (number of chronic conditions), $gender$, $school$, and $insurance$. 

## Estimation
```{r}
eq             = visits~hospital+health+chronic+gender+
                  school+insurance
bhat_pois      = glm(eq,data=NMES1988,family=poisson)
bhat_nb        = glm(eq,data=NMES1988)
bhat_hurdle    = hurdle(eq,data=NMES1988,dist="negbin")
bhat_zi        = zeroinfl(eq,data=NMES1988)
```

## Results
\tiny
```{r,echo=FALSE,warning=FALSE,message=FALSE}
stargazer(bhat_pois,bhat_nb,bhat_hurdle,bhat_zi,type="text",no.space=TRUE,omit.stat=c("ll","aic"))
```
\normalsize

# Survival Models
## Overview
Length of time until a certain event occurs and variables influencing time passed (also known as time-to-event data analysis). Examples:

- Time to failure of mechanical device
- Time to death after diagnosis with a certain disease
- Time to re-arrest after release from prison
- Time to defaulting on loan or mortgage

Data used for this topic

- `rossi`

## Theoretical Aspects
$T$ as a random variable representing survival time with the cumulative distribution function written as: 
$$F(t)=Pr(T \leq t)$$
where $t$ is a realization of $T$. Survival function as the complement probability (at least $t$)  :
$$S(t)=1-F(t)=Pr(T \geq t)$$
Hazard function or hazard rate $h(t)$ as risk of failure at time $t$.

## Example Data `rossi`
Experimental recidivism study on 432 male prisoners over a period of one year after release from prison ([Rossi et al., 1980](https://doi.org/10.1016/C2013-0-11412-2)):

- $week$: Week of first arrest after release
- $arrest$: Event indicator equal to 1 for rearrest during study period
- $fin$: Receipt of financial aid after release from prison (randomly assigned factor by the researchers)
- $age$: Age at the time of release
- $race$: Black and other
- $wexp$: Full-time work experience prior to incarceration
- $mar$: Married at the time of release
- $paro$: Released on parole
- $prio$: Number of prior convictions.
- $educ$: Education coded as 2 (grade 6 or less), 3 (grades 6-9), 4 (grades 10-11), 5 (grade 12), or 6 (some post-secondary).

## Analysis
Number of prisoners rearrested during study period:
```{r}
sum(rossi$arrest)
```

Surival object in R created by function `Surv()`:
```{r,results=FALSE,fig.keep='none'}
bhatmar = survfit(Surv(week,arrest)~mar,data=rossi)
bhatfin = survfit(Surv(week,arrest)~fin,data=rossi)
ggsurvplot(bhatmar,pval=TRUE,risk.table=TRUE)
ggsurvplot(bhatfin,pval=TRUE,risk.table=TRUE)
```

## Survival Curve: Marriage
```{r,echo=FALSE,fig.width=10,fig.height=5}
ggsurvplot(bhatmar,pval=TRUE,conf.int=TRUE,risk.table=TRUE)
```

## Survival Curve: Financial Aid
```{r,echo=FALSE,fig.width=10,fig.height=5}
ggsurvplot(bhatfin,pval=TRUE,conf.int=TRUE,risk.table=TRUE)
```

## Cumulative Hazard Function: Marriage
```{r,echo=FALSE,fig.width=10,fig.height=5}
ggsurvplot(bhatmar,pval=TRUE,conf.int=TRUE,fun="cumhaz")
```

## Cumulative Hazard Function: Financial Aid
```{r,echo=FALSE,fig.width=10,fig.height=5}
ggsurvplot(bhatfin,pval=TRUE,conf.int=TRUE,fun="cumhaz")
```

## Cox Regression in R: Setup
```{r}
bhat1 = coxph(Surv(week,arrest)~mar,data=rossi)
bhat2 = coxph(Surv(week,arrest)~mar+fin,data=rossi)
bhat3 = coxph(Surv(week,arrest)~fin+age+race+wexp+mar+paro+prio,
              data=rossi)
```

Statistically insignificant variables excluded from regression output on next slide due to space constraints: $paroyes$, $raceother$, and $wexpyes$

- In general, all variables must be reported! 

## Cox Regression in R: Results
\tiny
```{r,echo=FALSE}
stargazer(bhat1,bhat2,bhat3,type="text",no.space=TRUE,omit=c("paroyes","raceother","wexpyes"))
```
\normalsize

```{r,echo=FALSE}
knitr::knit_exit()
```



## Exercise
Consider the following data set:
```{r}
data(cancer,package="survival")
```
A description of the data can be found [here]()





```{r}
res.cox <- coxph(Surv(time, status) ~ age + sex + ph.ecog, data =  lung)
summary(res.cox)
```


http://www.sthda.com/english/wiki/cox-proportional-hazards-model
https://bioconnector.github.io/workshops/r-survival.html#survival_curves
https://rstudio-pubs-static.s3.amazonaws.com/293085_4739f7ad78114ce19330494a57366f86.html