---
title: "Panel Data"
author: "Jerome Dumortier"
date: "`r format(Sys.time(),'%d %B %Y')`"
output: 
     beamer_presentation:
          theme: "Hannover"
classoption: "aspectratio=169"
linkcolor: MidnightBlue
---

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
load("G:/My Drive/Teaching/Data Analysis for Public Affairs/GitHub/DataAnalysisPAData.RData")
library(stargazer)
```

## Overview

Required R Packages

- [plm](https://cran.r-project.org/web/packages/plm/index.html)
- [lmtest](https://cran.r-project.org/web/packages/lmtest/index.html)

    Documentation:
        
            - \url{https://cran.r-project.org/web/packages/plm/vignettes/plm.pdf}
        
## Note regarding Notation
Use of the package [stargazer](https://cran.r-project.org/web/packages/stargazer/index.html) due to the large number of variables. Slight change in notation:


##Introduction I} %For the NLSY79: Accessing data, investigator, continue as guest, pick income
    Pooled data: Combination of multiple cross-sectional data over time
        
            - Two or more different observational units over time
            - Grades in an economics class based on students' concentration combined from multiple semesters
            - American Community Survey (ACS): \url{http://www.census.gov/acs/www/}
        
    Panel data: Repeated measurement on the same individuals $i$ over time $t$.
        
            - Individual units can be people, states, firms, counties, countries, etc.
            - National Longitudinal Survey (NLSY79): \url{https://www.nlsinfo.org/content/getting-started}
            - Necessary adjustments of standard error due to correlation across time.
        
## Introduction II
Some assumptions about linear panel models:
        
- Regular time intervals
- Errors are correlated
- Parameters may vary across individuals or time
- Intercept: Individual specific effects model (fixed or random)
        
Note that the General Social Survey (GSS) is not a panel data set because different respondents are questioned every year.

## Examples and Advantages
Panel Study of Income Dynamics (PSID)

- Data on approximately 5,000 families on various socioeconomic and demographic variables

Survey of Income and Program Participation (SIPP)

- Interviews about economic condition of respondents

Advantages

- Takes into account heterogeneity among observational units, e.g., firms, states, counties, etc.
- Better understanding on the dynamics of change for observational units over time.
- Combines cross-sectional data with time series data leading to more complete behavioral models
        
## Terminology and Types
Balanced versus unbalance panel:

- A balanced panel has the same number of time-series observations for each subject or observational unit, whereas an unbalanced panel does not.
        
Short versus long panel:
        
- A short panel has a larger number of subjects or observational units than there are time periods. A long panel has a greater number of time periods than observational units.
        
Types of regression models:
        
- Pooled Ordinary Least Square model
- Fixed effects model
- Random effects model

# Simple Panel Data Methods
## Data `fertil1`: Variables
Data from the General Social Survey for the years 1974 to 1984
        
- $year$: 72 to 84, even
- $educ$:  years of schooling
- $meduc$ and $feduc$: mother's and father's education
- $kids$: number children ever born
- $east$, $northcentral$, and $west$: 1 if lived in at 16
- $far$m: 1 if on farm at 16
- $otherrural$: 1 if other rural at 16
- $town$: 1 if lived in town at 16
- $smallcity$: 1 if in small city at 16
        
Source: Jeffrey Wooldridge, Introductory Econometrics: A Modern Approach

##  Data Methods (`fertil1`)
```{r,echo=FALSE}
summary(lm(kids~educ+age+I(age^2)+east+northcentral+west+farm+otherrural
    +town+smallcity+y74+y76+y78+y80+y82+y84,data=fertil1))
```

\scriptsize \begin{verbatim}
lm(kids~educ+age+I(age^2)+east+northcentral+west+farm+otherrural
    +town+smallcity+y74+y76+y78+y80+y82+y84)

             Estimate Std. Error t value Pr(>|t|)
(Intercept) -6.785069   3.098715  -2.190 0.028758 *
educ        -0.129765   0.018653  -6.957 5.94e-12 ***
age          0.499186   0.140592   3.551 0.000400 ***
I(age^2)    -0.005436   0.001589  -3.421 0.000648 ***
east         0.060729   0.132538   0.458 0.646896
northcentral 0.219568   0.120638   1.820 0.069019 .
west         0.050807   0.167982   0.302 0.762360
[...] (Not statistically significant)
y82         -0.414209   0.174412  -2.375 0.017723 *
y84         -0.565326   0.177398  -3.187 0.001479 **
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 1.581 on 1112 degrees of freedom
Multiple R-squared:  0.09941,	Adjusted R-squared:  0.08645
F-statistic: 7.671 on 16 and 1112 DF,  p-value: < 2.2e-16
\end{verbatim}

## Interpretation
Evolution of fertility rates over time after controlling of other observable factors:
        
- Base year: 1972
- Negative coefficients indicate a drop in fertility in the early 1980's
- Coefficient of \emph{y82} (-0.41) indicates that women had on average 0.41 less children, i.e., 100 women had 41 kids less than 1972
- This drop is independent from education since we are controlling for education.
- More educated women have fewer children
- Assumes that the effect of each explanatory variable remains constant.
        
## Example using `cps7885`
Interact year dummy with key explanatory variables to see if the effect of that variable has changed over time:
$$\ln(wage)=\beta_0 + \gamma_0 \cdot y85 + \beta_1 \cdot educ + \gamma_1 \cdot y85 \cdot educ + \beta_2 \cdot exper\\ + \beta_3 \cdot exper^2 + \beta_4 \cdot union +\beta_5 \cdot female + \gamma_5 \cdot y85 \cdot female$$
Interpretation:
        
- $\beta_0$ is the 1978 intercept
- $\beta_0+\gamma_0$ is the 1985 intercept
- $\beta_1$ is the return to education in 1978
- $\beta_1 + \gamma_1$ is the return to education in 1985
- $\gamma_1$ measures how the return to education has changed over the seven year period
        
## Results for \texttt{cps7885}
```{r}
bhat=lm(log(wage) ~ y85 + educ + y85 * educ + exper + expersq +
    union + female + y85fem, data = cps7885)
```

\scriptsize \begin{verbatim}


Coefficients:
              Estimate Std. Error t value Pr(>|t|)
(Intercept)  4.589e-01  9.345e-02   4.911 1.05e-06 ***
y85          1.178e-01  1.238e-01   0.952   0.3415
educ         7.472e-02  6.676e-03  11.192  < 2e-16 ***
exper        2.958e-02  3.567e-03   8.293 3.27e-16 ***
expersq     -3.994e-04  7.754e-05  -5.151 3.08e-07 ***
union        2.021e-01  3.029e-02   6.672 4.03e-11 ***
female      -3.167e-01  3.662e-02  -8.648  < 2e-16 ***
y85fem       8.505e-02  5.131e-02   1.658   0.0977 .
y85:educ     1.846e-02  9.354e-03   1.974   0.0487 *
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 0.4127 on 1075 degrees of freedom
Multiple R-squared:  0.4262,	Adjusted R-squared:  0.4219
F-statistic:  99.8 on 8 and 1075 DF,  p-value: < 2.2e-16
\end{verbatim}

%========================================================================================================================================
##Simple Panel Data Methods}{Example using \texttt{cps7885}}
    Interpretation
        
            - 1978 return to education: 7.47\%
            - 1985 return to education: 7.47\%+1.85\% = 9.32\%
            - 1978 gender gap: 31.67\%
            - 1985 gender gap: 31.67\% - 8.51\% = 23.16\%
        
## Data `kiel`: Setup
Data set about home values near the location of an garbage incinerator

- Run 1981 data
- Run 1978 data

Difference-in-difference estimator: -\$30,688-(-\$18,824)=-\$11,864
$$ \hat{\delta}_1 = (price_{81,near}-price_{81,far})-(price_{78,near}-price_{78,far})$$
where $\hat{\delta}_1$ represents the difference over time in average differences in housing prices in the two locations.

%========================================================================================================================================
##Simple Panel Data Methods}{Example using \texttt{kiel}}
    To determine statistical significance:
        \begin{equation*}
            price = \beta_0 + \gamma_0 \cdot y81 + \beta_1 \cdot nearinc + \gamma_1 \cdot y81 \cdot nearinc
        \end{equation*}
    Interpretation
        
            - $\beta_0$: Average home value which is not near the garbage incinerator
            - $\gamma_0 \cdot y81$: Average change in housing values for all homes
            - $\beta_1 \cdot nearinc$: Location effect that is not due to the incinerator
            - $\gamma_1$: Decline in housing values due to incinerator
        
    Homes have lost 9.3\% in values when including additional independent variables and using the natural logarithm of price.

%========================================================================================================================================
%##Two Period Panel Analysis}{Same Observational Unit for $t=1,2$ using \texttt{crime}}
%    Crime data on 92 cities for 1982 and 1987
%        
%            - Run \texttt{lm(crmrte~unem,data=crime87)}
%            - Unemployment lowers crime. Can this be correct?
%        
%    Unobserved factors affecting dependent variable:
%        \begin{enumerate}
%            - Constant over time
%            - Varying over time
%        \end{enumerate}
%
%========================================================================================================================================
%##Two Period Panel Analysis}{Same Observational Unit for $t=1,2$ using \texttt{crime}}
%    First regression suffers from omitted variable bias, i.e., views the unobserved factors as (1) constant and (2) varying over time
%        \begin{equation*}
%            y_{it} = \beta_0+ \gamma_0 \cdot d_t+\beta_1 \cdot x_{it} + a_i +\epsilon_{it}
%        \end{equation*}
%    where
%        
%            - $\beta_0$: $t=1$ intercept ($t=2$: $\beta_0+\gamma_0$)
%            - $d_t$: Dummy variable for time periods (equal to 1 for time period 2)
%            - $a_i$: unobserved time constant effect (fixed effect over time)
%        
%    This is an unobserved effects model.
%
%========================================================================================================================================
%##Unobserved Effects Model for Crime}
%        \begin{equation*}
%            crate = \beta_0 + \gamma_0 \cdot d87 + \beta_1 \cdot unemployment + a_i + \epsilon_{it}
%        \end{equation*}
%    where $a_i + \epsilon_{it}$ is a composite error. The composite error is a problem because it may be correlated with the independent variables. Solution:
%        \begin{align*}
%            y_{i2} & = (\beta_0+\gamma_0)+\beta_1 \cdot x_{i2} + a_i + \epsilon_{i2}\\
%            y_{i1} & = \beta_0+\beta_1 \cdot x_{i1} + a_i + \epsilon_{i1}
%        \end{align*}
%    Taking the difference between the two equations:
%        \begin{equation*}
%            y_{i2}-y_{i1} = \gamma_0+\beta_1 \cdot (x_{i2}-x_{i1}) + (\epsilon_{i2} - \epsilon_{i1})
%        \end{equation*}
%    This can be written as
%        \begin{equation*}
%            \Delta y = \gamma_0 + \beta_1 \cdot \Delta x + \Delta \epsilon
%        \end{equation*}
%    The last equation satisfies all the assumption of the classical model.
%
%========================================================================================================================================
##Grunfeld Data}
    The data set is used in many textbooks and comes with the package \texttt{plm}.
        
            - After loading the package, type \texttt{data(``Grunfeld'')}
        
    Data on 10 companies over the period 1935 to 1954:
        
            - \emph{inv}: Investment
            - \emph{value}: Value of the firm
            - \emph{capital}: Capital stock
        
    Companies of interest for this class: GM (firm 1), U.S. Steel (firm 2), GE (firm 3), Westinghouse (firm 8)

%========================================================================================================================================
##Pooled OLS Model}
    Pooling all cross-sectional and time series observations into a single data set and running an OLS regression.
        \begin{equation*}
            inv_i = \beta_0 + \beta_1 \cdot value_i + \beta_2 \cdot capital_i
        \end{equation*}
    General formulation of the pooled model
        \begin{equation*}
            y_{it}=\beta_0+\beta_1 \cdot x_i + \epsilon_i
        \end{equation*}
    Issues with pooled OLS model:
        
            - Ignores heterogeneity among the observations and time.
            - If heterogeneity is present\\
             $\Rightarrow$ Correlation between independent variables and error term leading to biased and inconsistent coefficient estimates. Fix: Fixed effects model takes heterogeneity into account.\\
             $\Rightarrow$ Autocorrelation between error terms. Fix: Random effects model
        

%========================================================================================================================================
\begin{frame}[fragile]{Data Preparation and Pooled OLS Model}
    We are only interested in firm 1, 2, 3, and 8:
        
            - \small \verb"grunfeld = subset(Grunfeld,Grunfeld$firm %in% c(1,2,3,8))"
        
    To use the functions from \texttt{plm}, define data as a panel data set:
        
            - \small \verb+grunfeld = pdata.frame(grunfeld,index=c("firm","year"))+
        
    Running a simple OLS model on the data:
        
            - Using the regular \texttt{lm} function
            - Using the \texttt{plm} function and specifying the model as \texttt{``pooling''}
            - Name the outputs \texttt{``grunwald.ols''} and \texttt{``grunwald.pooling''}
        

%========================================================================================================================================
\begin{frame}[fragile]{Pooled OLS Model}
\scriptsize \begin{verbatim}
Call:
lm(formula = inv ~ value + capital, data = grunfeld)

Residuals:
    Min      1Q  Median      3Q     Max
-319.68  -99.95    1.96   65.99  336.21

Coefficients:
             Estimate Std. Error t value Pr(>|t|)
(Intercept) -62.83184   29.72538  -2.114   0.0378 *
value         0.11052    0.01378   8.023 9.19e-12 ***
capital       0.30046    0.04940   6.082 4.27e-08 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 142.9 on 77 degrees of freedom
Multiple R-squared:  0.7547,	Adjusted R-squared:  0.7483
F-statistic: 118.4 on 2 and 77 DF,  p-value: < 2.2e-16
\end{verbatim}

# Fixed Effects Model
## Theoretical Concepts I
Fixed effects model or Least-Squares Dummy Variable (LSDV) regression
        
            - Constant slope coefficients but varying intercept over $i$
        
    Regression equation:
        \begin{gather*}
            inv_{it} = \beta_{0i} + \beta_1 \cdot value_{it} + \beta_2 \cdot capital_{it}\\
            i=1,2,3,4\\
            t=1,2, \dots 20
        \end{gather*}
    This model can also be written as
        \begin{equation*}
            inv_{it} = \alpha_0+\alpha_1 \cdot D_{1i} + \alpha_2 \cdot D_{2i} +\alpha_3 \cdot D_{3i} + \beta_1 \cdot value_{it} + \beta_2 \cdot capital_{it}
        \end{equation*}
    Individual specific effects:
        \begin{equation*}
            y_{it} = \alpha_i + \beta_i \cdot x_{it} + \epsilon_{it}
        \end{equation*}
    $\alpha_i$ can be fixed or random

%========================================================================================================================================
##Fixed Effects Model}{Theoretical Concepts II}
    Fixed effects model
        
            - Intercept $\beta_{0i}$ is firm specific.
            - For an individual, this could be education and/or ability, possibly correlated with independent variables
            - Intercept is time-invariant.
            - Slope coefficients do not vary across individuals (firms) or time
        
    Implementation in R
        
            - Using the \texttt{plm} function and specifying the model as \texttt{``within''}
            - Name the output \texttt{``grunwald.fixed''}
        

%========================================================================================================================================
\begin{frame}[fragile]{Fixed Effects Model}{Implementation in R using \texttt{plm}}
\scriptsize \begin{verbatim}
Call:
plm(formula = inv ~ value + capital, data = grunfeld, model = "within")

Balanced Panel: n=4, T=20, N=80

Coefficients :
        Estimate Std. Error t-value Pr(>|t|)
value   0.108400   0.017566  6.1711  3.3e-08 ***
capital 0.345058   0.026708 12.9195  < 2e-16 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Total Sum of Squares:    2171500
Residual Sum of Squares: 422220
R-Squared:      0.80556
Adj. R-Squared: 0.79242
F-statistic: 153.291 on 2 and 74 DF, p-value: < 2.22e-16
\end{verbatim}

%========================================================================================================================================
\begin{frame}[fragile]{Fixed Effects Model}{Firm-Specific Intercepts and Hypothesis Test}
In order to get the firm specific intercepts:
\begin{verbatim}
fixef(grunfeld.fixed)
         1          2          3          8
 -85.51533   94.98762 -246.22751  -59.38573
\end{verbatim}

Testing whether a fixed effects or OLS is appropriate (H$_0$: OLS better):
\begin{verbatim}
pFtest(grunfeld.fixed,grunfeld.ols)

	F test for individual effects

data:  inv ~ value + capital
F = 67.215, df1 = 3, df2 = 74, p-value < 2.2e-16
alternative hypothesis: significant effects
\end{verbatim}

If the p-value is below 0.05 then the fixed effects model is a better choice.

%========================================================================================================================================
\begin{frame}[fragile]{Fixed Effects Model}{Implementation in R using \texttt{lm}}
\scriptsize \begin{verbatim}
Call:
lm(formula = inv ~ value + capital + factor(firm), data = grunfeld)

Coefficients:
                Estimate Std. Error t value Pr(>|t|)
(Intercept)    -85.51533   73.48978  -1.164 0.248310
value            0.10840    0.01757   6.171  3.3e-08 ***
capital          0.34506    0.02671  12.919  < 2e-16 ***
factor(firm)2  180.50295   45.71679   3.948 0.000178 ***
factor(firm)3 -160.71218   46.62236  -3.447 0.000938 ***
factor(firm)8   26.12959   64.94348   0.402 0.688591
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 75.54 on 74 degrees of freedom
Multiple R-squared:  0.9341,	Adjusted R-squared:  0.9297
F-statistic: 209.9 on 5 and 74 DF,  p-value: < 2.2e-16
\end{verbatim}

%========================================================================================================================================
##Random Effects Model}{Theoretical Concepts}
    The general fixed effects model can be expressed as
        \begin{equation*}
            y_{it} = \beta_{0i} + \beta_1 \cdot x_{1,it} + \beta_2 \cdot x_{2,it} + \epsilon_{it}
        \end{equation*}
    Instead of treating $\beta_{0i}$ as fixed, the random model assumes
        \begin{equation*}
            \beta_{0i} = \beta_0 + \upsilon_i
        \end{equation*}
    where $\upsilon_i$ is random error term with a mean of zero and variance $\sigma_{\upsilon}^2$. According to Gujarati ``What we are essentially saying is that the four firms included in our sample are a drawing from a much larger universe of such companies and that they have a common mean value for the intercept $\beta_0$ and the individual differences in the intercept values of each company are reflected in the error term $\upsilon_i$.''

%========================================================================================================================================
\begin{frame}[fragile]{Random Effects Model}
\scriptsize \begin{verbatim}
Call: plm(formula = inv ~ value + capital, data = grunfeld, model = "random")

Effects:
                   var  std.dev share
idiosyncratic  5705.63    75.54 0.212
individual    21217.03   145.66 0.788
theta:  0.8848

Coefficients :
              Estimate Std. Error t-value  Pr(>|t|)
(Intercept) -73.084676  81.172222 -0.9004    0.3707
value         0.108056   0.016840  6.4165 1.034e-08 ***
capital       0.344543   0.026629 12.9384 < 2.2e-16 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Total Sum of Squares:    2227700
Residual Sum of Squares: 437750
R-Squared:      0.8035
Adj. R-Squared: 0.79839
F-statistic: 157.425 on 2 and 77 DF, p-value: < 2.22e-16
\end{verbatim}

%========================================================================================================================================
\begin{frame}[fragile]{Breusch-Pagan Lagrange Multiplier (LM) Test}
For random effects models: Null hypothesis of no panel effect, i.e., OLS is better. If p-value is below 0.05, we reject the null hypothesis and thus, a random effects model is more appropriate than the OLS.

\small \begin{verbatim}
plmtest(grunfeld.pooling,type=c("bp"))

	Lagrange Multiplier Test - (Breusch-Pagan) for balanced panels

data:  inv ~ value + capital
chisq = 378.44, df = 1, p-value < 2.2e-16
alternative hypothesis: significant effects
\end{verbatim}

%========================================================================================================================================
\begin{frame}[fragile]{Hausman Test: Fixed or Random Model}
The Hausman Test tests the null hypothesis that the preferred model is a random effects model. It basically tests whether the unique errors are correlated with the regressors.

\begin{verbatim}
phtest(grunfeld.random,grunfeld.fixed)

	Hausman Test

data:  inv ~ value + capital
chisq = 0.14882, df = 2, p-value = 0.9283
alternative hypothesis: one model is inconsistent
\end{verbatim}
If p-value is below 0.05 then use fixed effects.

%========================================================================================================================================
%\begin{frame}[fragile]{Testing for Heteroscedasticity}
%\begin{verbatim}
%bptest(inv~value+capital+factor(firm),data=grunfeld)
%
%	studentized Breusch-Pagan test
%
%data:  inv ~ value + capital + factor(firm)
%BP = 25.375, df = 5, p-value = 0.0001179
%\end{verbatim}
%If the p-value is below 0.05, then we face heteroscedasticity.
%
%========================================================================================================================================
%\begin{frame}[fragile]{Heteroscedasticity Consistent Coefficients and Standard Errors}
%\scriptsize \begin{verbatim}
%coeftest(grunfeld.fixed,vcovHC)
%
%t test of coefficients:
%
%        Estimate Std. Error t value  Pr(>|t|)
%value   0.108400   0.014293  7.5839 7.902e-11 ***
%capital 0.345058   0.031152 11.0765 < 2.2e-16 ***
%---
%Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
%
%coeftest(grunfeld.random,vcovHC)
%
%t test of coefficients:
%
%              Estimate Std. Error t value  Pr(>|t|)
%(Intercept) -73.084676  47.566742 -1.5365    0.1285
%value         0.108056   0.012149  8.8945 1.915e-13 ***
%capital       0.344543   0.032276 10.6750 < 2.2e-16 ***
%---
%Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
%\end{verbatim}
%








\end{document}

##Implementation in R}

    Data set: \emph{psid.csv} (Panel Study on Income Dynamics)
        
            - exp
            - wks
            - occ
            - ind
            - south
            - smsa
            - ms
            - fem
        



