---
title: "Panel Data"
author: "Jerome Dumortier"
date: "`r format(Sys.time(),'%d %B %Y')`"
output: 
     beamer_presentation:
          theme: "Hannover"
          slide_level: 3
section-titles: false
classoption: "aspectratio=169"
linkcolor: MidnightBlue
---

```{r,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE}
load("G:/My Drive/Teaching/Data Analysis for Public Affairs/GitHub/DataAnalysisPAData.RData")
library(lmtest)
library(plm)
library(stargazer)
```

## Overview
Required R Packages

- [plm](https://cran.r-project.org/web/packages/plm/index.html)
- [lmtest](https://cran.r-project.org/web/packages/lmtest/index.html)

Documentation
        
- [Panel data econometrics in R](https://cran.r-project.org/web/packages/plm/vignettes/A_plmPackage.html)
        
Note regarding Notation

- Use of the package [stargazer](https://cran.r-project.org/web/packages/stargazer/index.html) due to the large number of variables. 

## Introduction I
Pooled data: Combination of multiple cross-sectional data over time
        
- Two or more different observational units over time
- Grades in an economics class based on students' concentration combined from multiple semesters
- [American Community Survey (ACS)](http://www.census.gov/acs/www/)
        
 Panel data: Repeated measurement on the same individuals $i$ over time $t$.
        
- Individual units can be people, states, firms, counties, countries, etc.
- [National Longitudinal Survey (NLSY79)](https://www.nlsinfo.org/content/getting-started)
- Necessary adjustments of standard error due to correlation across time.

For NLSY79: Accessing data $\Rightarrow$ investigator $\Rightarrow$ Begin searching as guest $\Rightarrow$ Pick income as an example
        
## Introduction II
Some assumptions about linear panel models:
        
- Regular time intervals
- Errors are correlated
- Parameters may vary across individuals or time
- Intercept: Individual specific effects model (fixed or random)
        
Note that the General Social Survey (GSS) is not a panel data set because different respondents are questioned every year.

## Examples and Advantages
Panel Study of Income Dynamics (PSID)

- Data on approximately 5,000 families on various socioeconomic and demographic variables

Survey of Income and Program Participation (SIPP)

- Interviews about economic condition of respondents

Advantages

- Takes into account heterogeneity among observational units, e.g., firms, states, counties, etc.
- Better understanding on the dynamics of change for observational units over time.
- Combines cross-sectional data with time series data leading to more complete behavioral models
        
## Terminology and Types
Balanced versus unbalance panel:

- A balanced panel has the same number of time-series observations for each subject or observational unit, whereas an unbalanced panel does not.
        
Short versus long panel:
        
- A short panel has a larger number of subjects or observational units than there are time periods. A long panel has a greater number of time periods than observational units.
        
Types of regression models:
        
- Pooled Ordinary Least Square model
- Fixed effects model
- Random effects model

# Simple Panel Data Methods
## `fertil1`: Variables
Data from the General Social Survey for the years 1974 to 1984
        
- $year$: 72 to 84, even
- $educ$:  years of schooling
- $meduc$ and $feduc$: mother's and father's education
- $kids$: number children ever born
- $east$, $northcentral$, and $west$: 1 if lived in at 16
- $farm$: 1 if on farm at 16
- $otherrural$: 1 if other rural at 16
- $town$: 1 if lived in town at 16
- $smallcity$: 1 if in small city at 16
        
Source: Jeffrey Wooldridge, Introductory Econometrics: A Modern Approach

## `fertil1`: Estimation
\tiny
```{r,echo=FALSE,warning=FALSE}
bhat = lm(kids~educ+age+I(age^2)+east+northcentral+west+farm+otherrural+
          town+smallcity+y74+y76+y78+y80+y82+y84,data=fertil1)
stargazer(bhat,type="text",no.space=TRUE,single.row=TRUE,
          omit=c("farm","otherrural","town","smallcity","y74","y76","y78","y80"))
```
\normalsize

## `fertil1`: Interpretation
Evolution of fertility rates over time after controlling of other observable factors:
        
- Base year: 1972
- Negative coefficients indicate a drop in fertility in the early 1980's
- Coefficient of $y82$ (-0.41) indicates that women had on average 0.41 less children, i.e., 100 women had 41 kids less than 1972
- This drop is independent from education since we are controlling for education.
- More educated women have fewer children
- Assumes that the effect of each explanatory variable remains constant.
        
## `cps7885`: Setup 
Interact year dummy with key explanatory variables to see if the effect of that variable has changed over time:
$$
\begin{aligned}
\ln(wage)= &\beta_0 + \gamma_0 \cdot y85 + \beta_1 \cdot educ + \gamma_1 \cdot y85 \cdot educ + \beta_2 \cdot exper + \beta_3 \cdot exper^2\\  
 & + \beta_4 \cdot union +\beta_5 \cdot female + \gamma_5 \cdot y85 \cdot female
 \end{aligned}
$$
Interpretation:
        
- $\beta_0$ is the 1978 intercept
- $\beta_0+\gamma_0$ is the 1985 intercept
- $\beta_1$ is the return to education in 1978
- $\beta_1 + \gamma_1$ is the return to education in 1985
- $\gamma_1$ measures how the return to education has changed over the seven year period
        
## `cps7885`: Estimation
```{r,results='hide'}
cps7885$y85    = ifelse(cps7885$year==85,1,0)
bhat           = lm(log(wage)~y85+educ+y85:educ+exper+
                    I(exper^2)+union+female+y85:female,
                    data=cps7885)
```

## `cps7885`: Results
\tiny
```{r}
stargazer(bhat,type="text",no.space=TRUE,single.row=TRUE)
```
\normalsize

## `cps7885`: Interpretation
Interpretation
        
- 1978 return to education: 7.47\%
- 1985 return to education: 7.47\%+1.85\% = 9.32\%
- 1978 gender gap: 31.67\%
- 1985 gender gap: 31.67\% - 8.51\% = 23.16\%
        
## Data `kiel`: Setup I
Data set about home values near the location of an garbage incinerator

- Run 1981 data
- Run 1978 data

Difference-in-difference estimator: -\$30,688-(-\$18,824)=-\$11,864
$$ \hat{\delta}_1 = (price_{81,near}-price_{81,far})-(price_{78,near}-price_{78,far})$$
where $\hat{\delta}_1$ represents the difference over time in average differences in housing prices in the two locations.

## Data `kiel`: Setup II
To determine statistical significance:
$$price = \beta_0 + \gamma_0 \cdot y81 + \beta_1 \cdot nearinc + \gamma_1 \cdot y81 \cdot nearinc$$
Interpretation

- $\beta_0$: Average home value which is not near the garbage incinerator
- $\gamma_0 \cdot y81$: Average change in housing values for all homes
- $\beta_1 \cdot nearinc$: Location effect that is not due to the incinerator
- $\gamma_1$: Decline in housing values due to incinerator

Homes have lost 9.3\% in values when including additional independent variables and using the natural logarithm of price.

## Data `kiel`: Naive Implementation in R
```{r}
kiel81         = subset(kiel,year==1981)
bhat81         = lm(rprice~nearinc,data=kiel81)
kiel78         = subset(kiel,year==1978)
bhat78         = lm(rprice~nearinc,data=kiel78)
```

## Data `kiel`: Naive Results
\tiny
```{r,echo=FALSE,warning=FALSE}
stargazer(bhat81,bhat78,type="text",no.space=TRUE,single.row=TRUE)
```
\normalsize 

## Data `kiel`: Implementation in R
```{r}
bhat1     = lm(rprice~y81+nearinc+y81:nearinc,data=kiel)
bhat2     = lm(rprice~y81+nearinc+y81:nearinc+age+agesq,data=kiel)
bhat3     = lm(rprice~y81+nearinc+y81:nearinc+age+agesq+cbd+rooms+
               area+land+baths,data=kiel)
```

## Data `kiel`: Results
\tiny
```{r,echo=FALSE,warning=FALSE}
stargazer(bhat1,bhat2,bhat3,type="text",no.space=TRUE,
          omit=c("age","agesq","cbd","bath"))
```

## Grunfeld Data
The data set is used in many textbooks and comes also with the package [plm](https://cran.r-project.org/web/packages/plm/index.html). Data on 10 companies over the period 1935 to 1954:

- $inv$: Investment
- $value$: Value of the firm
- $capital$: Capital stock

Companies of interest for this class: GM (firm 1), U.S. Steel (firm 2), GE (firm 3), Westinghouse (firm 8)

## Pooled OLS Model
Pooling all cross-sectional and time series observations into a single data set and running an OLS regression.
$$inv_i = \beta_0 + \beta_1 \cdot value_i + \beta_2 \cdot capital_i$$
General formulation of the pooled model
$$y_{it}=\beta_0+\beta_1 \cdot x_i + \epsilon_i$$
Issues with pooled OLS model:

- Ignores heterogeneity among the observations and time.

With heterogeneity: Biased and inconsistent estimates due to correlation between independent variables and error term. 

## Data Preparation and Pooled OLS Model
Use of firms 1, 2, 3, and 8:
```{r}
grunfeld = subset(grunfeld,grunfeld$firm %in% c(1,2,3,8))
```
To use the functions from [plm](https://cran.r-project.org/web/packages/plm/index.html), define data as a panel data set:
```{r}
grunfeld = pdata.frame(grunfeld,index=c("firm","year"))
```
Running a simple OLS model on the data:
        
- Using the regular `lm()` function
- Using the `plm()` function and specifying the model as **pooling**
- Name the outputs `grunfeld.ols` and `grunfeld.pooling`
        
## Pooled OLS Model
```{r}
grunfeld.ols       = lm(inv~value+capital,data=grunfeld)
grunfeld.pooling   = plm(inv~value+capital,data=grunfeld,
                         model="pooling")
```

## Pooled OLS Model
\tiny
```{r,warning=FALSE}
stargazer(grunfeld.ols,grunfeld.pooling,no.space=TRUE,
          single.row=TRUE,type="text")
```
\normalsize

# Fixed Effects Model
## Theoretical Concepts I
Fixed effects model or Least-Squares Dummy Variable (LSDV) regression
        
- Constant slope coefficients but varying intercept over $i$
        
Regression equation:
$$inv_{it} = \beta_{0i} + \beta_1 \cdot value_{it} + \beta_2 \cdot capital_{it}$$
with $i=1,2,3,4$ and $t=1,2, \dots, 20$. This model can also be written as
$$inv_{it} = \alpha_0+\alpha_1 \cdot D_{1i} + \alpha_2 \cdot D_{2i} +\alpha_3 \cdot D_{3i} + \beta_1 \cdot value_{it} + \beta_2 \cdot capital_{it}$$
Individual specific effects:
$$y_{it} = \alpha_i + \beta_i \cdot x_{it} + \epsilon_{it}$$
$\alpha_i$ can be fixed or random

## Theoretical Concepts II
Fixed effects model

- Intercept $\beta_{0i}$ is firm specific.
- For an individual, this could be education and/or ability, possibly correlated with independent variables
- Intercept is time-invariant.
- Slope coefficients do not vary across individuals (firms) or time

Implementation in R
```{r}
grunfeld.fixed = plm(inv~value+capital,data=grunfeld,
                     model="within")
```

## Results
\tiny
```{r}
stargazer(grunfeld.fixed,no.space=TRUE,
          single.row=TRUE,type="text")
```
\normalsize

## Firm-Specific Intercepts and Hypothesis Test
In order to get the firm specific intercepts:

\tiny
```{r}
fixef(grunfeld.fixed)
```
\normalsize

Testing whether a fixed effects or OLS is appropriate (H$_0$: OLS better):

\tiny
```{r}
pFtest(grunfeld.fixed,grunfeld.ols)
```
\normalsize

If the p-value is below 0.05 then the fixed effects model is a better choice.

## Implementation in R using `lm()`
Implementation of a fixed effects model with the command `lm()`
```{r}
bhat = lm(inv~value+capital+factor(firm),data=grunfeld)
```

# Random Effects Model
## Theoretical Concepts
The general fixed effects model can be expressed as
$$y_{it} = \beta_{0i} + \beta_1 \cdot x_{1,it} + \beta_2 \cdot x_{2,it} + \epsilon_{it}$$
Instead of treating $\beta_{0i}$ as fixed, the random model assumes
$$\beta_{0i} = \beta_0 + \upsilon_i$$
where $\upsilon_i$ is random error term with a mean of zero and variance $\sigma_{\upsilon}^2$. According to Gujarati: *What we are essentially saying is that the four firms included in our sample are a drawing from a much larger universe of such companies and that they have a common mean value for the intercept $\beta_0$ and the individual differences in the intercept values of each company are reflected in the error term $\upsilon_i$.*

## Implementation in R
\tiny
```{r}
grunfeld.random = plm(inv~value+capital,data=grunfeld,model="random")
stargazer(grunfeld.random,no.space=TRUE,single.row=TRUE,type="text")
```
\normalsize

## Breusch-Pagan Lagrange Multiplier (LM) Test
For random effects models: Null hypothesis of no panel effect, i.e., OLS is better. If p-value is below 0.05, we reject the null hypothesis and thus, a random effects model is more appropriate than the OLS.

```{r}
plmtest(grunfeld.pooling,type=c("bp"))
```

## Hausman Test: Fixed or Random Model
The Hausman Test tests the null hypothesis that the preferred model is a random effects model. It basically tests whether the unique errors are correlated with the regressors.

```{r}
phtest(grunfeld.random,grunfeld.fixed)
```

## Testing for Heteroscedasticity
```{r}
bptest(inv~value+capital+factor(firm),data=grunfeld)
```

If the p-value is below 0.05, then we have heteroscedasticity.

## Heteroscedasticity Consistent Coefficients and Standard Errors
```{r}
coeftest(grunfeld.fixed,vcovHC)
```
