---
title: "Introduction to Probability Distribution Fitting"
author: "Jerome Dumortier"
date: "`r format(Sys.time(),'%d %B %Y')`"
output: 
     beamer_presentation:
          theme: "Hannover"
classoption: "aspectratio=169"
linkcolor: MidnightBlue
---

```{r,echo=FALSE,warning=FALSE,results=FALSE,message=FALSE}
load("G:/My Drive/Teaching/Data Analysis for Public Affairs/GitHub/DataAnalysisPAData.RData")
library(fitdistrplus)
library(gamlss)
library(ggplot2)
library(invgamma)
library(MASS)
```

## Lecture Overview
Distribution fitting

- Finding the best-fitting theoretical probability distribution for the observed data

Three approaches covered in this lecture:

- [MASS](https://cran.r-project.org/web/packages/MASS/index.html): `fitdistr()` 
- [fitdistrplus](https://cran.r-project.org/web/packages/fitdistrplus/index.html): `fitdist()` 
- [gamlss](https://cran.r-project.org/web/packages/gamlss/index.html): `fitDist()` 

Notes:

- No need to specify distribution function for the last approach, i.e., `fitDist()`
- Introduction and overview to a very broad field of research

## Introduction
Empirical work often requires understanding of the underlying distribution of data:

- Distribution of corn yields in a particular county based on observations to calculate the probability of getting a yield below a certain threshold, e.g., for crop insurance purposes 
- Wind speed distribution at a particular location for construction of a wind farm: Electricity production is not possible below and above a certain wind speed

Estimation of one or more parameters characterizing a probability distribution function

# Introductory Example 

## Weibull: Random Data Generation
Random generation of data (N=10000) following a Weibull distribution with two parameters:

- Shape: $k=2$ 
- Scale: $\lambda=1.5$

```{r}
weibulldata = rweibull(10000,2,1.5)
```

## Weibull: Histogram
```{r,fig.height=3,fig.width=5,echo=FALSE}
hist(weibulldata,main="Histogram of Weibull Data")
```

## Weibull: Distribution Fitting with `fitdistr`
```{r}
weibullpara    = fitdistr(weibulldata,densfun="weibull",
                          lower=c(0,0))
shape          = weibullpara$estimate[1]
scale          = weibullpara$estimate[2]
c(shape,scale)
```

## Weibull: Observed Data and Estimated Distribution
```{r,fig.height=2.5,fig.width=4,fig.align='center'}
hist(weibulldata,freq=FALSE,ylim=c(0,0.6),xlim=c(0,6))
range          = seq(0,6,0.1)
lines(range,dweibull(range,shape,scale))
```

# Approach

## Distribution Fitting Steps
General steps (see [Fitting Distributions with R](https://cran.r-project.org/doc/contrib/Ricci-distributions-en.pdf) by Vito Ricci for more information)

1. General hypothesis about candidate distributions, e.g., discrete vs. continuous, entire real number line vs. positive numbers only
     - Histogram as a valuable first approach
2. Parameter estimation
     - Example: Calculating shape and scale parameters of the Weibull distribution or mean and variance for a Normal distribution 
3. Goodness of fit

Starting point for an overview of various probability distributions: [List of probability distributions](https://en.wikipedia.org/wiki/List_of_probability_distributions)

# Candidate Distributions and Estimation

## Meridian Hills: Possible Distributions
Meridian Hills home values:

- Source: \url{https://jrfdumortier.github.io/dataanalysis/}
- 101 home values in the Meridian Hills neighborhood in Indianapolis
- Scaling of data to measure home values in \$1000 

Candidate distributions: 

- Gamma distribution: Shape and scale parameter
- Weibull distribution: Shape and scale parameter
- Log-normal distribution, i.e, $Y=\ln(X)$ has a normal distribution: $\mu$ and $\sigma$

```{r}
mhprice        = mh1$price/1000
mhgamma        = fitdistr(mhprice,"gamma")
mhweibull      = fitdistr(mhprice,"weibull",lower=c(0,0))
mhlognormal    = fitdistr(mhprice,"log-normal")
```

## Meridian Hills: Histogram I
```{r,fig.keep='none'}
hist(mhprice,freq=FALSE,ylim=c(0,0.0025),
     xlim=c(0,2000),main="Meridian Hills")
range          = seq(0,2000,1)
lines(range,dgamma(range,mhgamma$estimate[1],
                   mhgamma$estimate[2]),col="blue")
lines(range,dweibull(range,mhweibull$estimate[1],
                     mhweibull$estimate[2]),col="red")
lines(range,dlnorm(range,mhlognormal$estimate[1],
                   mhlognormal$estimate[2]),col="green")
```

## Meridian Hills: Histogram II
```{r,echo=FALSE,fig.height=3,fig.width=5,fig.align='center'}
hist(mhprice,freq=FALSE,ylim=c(0,0.0025),
     xlim=c(0,2000),main="Meridian Hills")
range          = seq(0,2000,1)
lines(range,dgamma(range,mhgamma$estimate[1],
                   mhgamma$estimate[2]),col="blue")
lines(range,dweibull(range,mhweibull$estimate[1],
                     mhweibull$estimate[2]),col="red")
lines(range,dlnorm(range,mhlognormal$estimate[1],
                   mhlognormal$estimate[2]),col="green")
```

# Goodness of Fit

## Meridian Hills: Setup for `fitdist()`
Use of the function `fitdist()` from the package [fitdistrplus](https://cran.r-project.org/web/packages/fitdistrplus/index.html)

```{r,warning=FALSE}
mhprice        = mh1$price/1000
mhgamma        = fitdist(mhprice,"gamma",lower=c(0,0))
mhweibull      = fitdist(mhprice,"weibull",lower=c(0,0))
mhlognormal    = fitdist(mhprice,"lnorm",lower=c(0,0))
```

## Meridian Hills: Gamma Distribution
```{r,echo=FALSE,fig.height=5.5,fig.align='center'}
plot(mhgamma)
```

## Meridian Hills: Weibull Distribution
```{r,echo=FALSE,fig.height=5.5,fig.align='center'}
plot(mhweibull)
```

## Meridian Hills: Log-Normal Distribution
```{r,echo=FALSE,fig.height=5.5,fig.align='center'}
plot(mhlognormal)
```

## Ground Beef: Possible Distributions
Second example using the function `fitdist()` package:

- Use of the data `groundbeef` associated with the package `fitdistrplus`: Serving sizes collected in a French survey, for ground beef patties consumed by children under 5 years old.

```{r}
data("groundbeef")
gbnormal       = fitdist(groundbeef$serving,"norm")
gbweibull      = fitdist(groundbeef$serving,"weibull")
gbgamma        = fitdist(groundbeef$serving,"gamma")
gblognormal    = fitdist(groundbeef$serving,"lnorm")
fitteddist     = list(gbnormal,gbweibull,gbgamma,gblognormal)
plotlegend     = c("Normal","Weibull","Gamma","Lognormal")
```

## Ground Beef: Histogram
```{r,echo=FALSE,fig.height=3,fig.width=5,fig.align='center'}
hist(groundbeef$serving,main="Ground Beef",xlab="Data",xlim=c(0,200))
```

## Ground Beef:  Results I
```{r,echo=FALSE,fig.height=3,fig.width=5,fig.align='center'}
denscomp(fitteddist,legendtext=plotlegend)
```

## Ground Beef:  Results II
```{r,echo=FALSE,fig.height=3,fig.width=5,fig.align='center'}
cdfcomp(fitteddist,legendtext=plotlegend)
```

## Results: Q-Q Plot
```{r,echo=FALSE,fig.height=3,fig.width=5,fig.align='center'}
qqcomp(fitteddist,legendtext=plotlegend)
```

## Unspecified Distribution: `fitDist()`
Use of the function `fitDist()` from package [gamlss](https://cran.r-project.org/web/packages/gamlss/index.html)

```{r,warning=FALSE,message=FALSE,comment=FALSE,error=FALSE,results='hide'}
output = fitDist(mhprice,type="realplus")
```

```{r}
output$family
output$Allpar
```

## Goodness of Fit with Inverse Gamma
```{r,echo=FALSE,fig.height=5.5,fig.align='center'}
plot(output,summaries=FALSE)
```

# Discrete Data Distribution Fitting

## EV Data
```{r,warning=FALSE,message=FALSE,comment=FALSE,error=FALSE,results='hide'}
evpoisson = fitdist(evdata$numcars,discrete=TRUE,distr="pois")
evnbinom  = fitdist(evdata$numcars,discrete=TRUE,distr="nbinom")
```

## EV Data: Results Poisson
```{r,echo=FALSE,fig.height=3,fig.width=5,fig.align='center'}
plot(evpoisson)
```

## EV Data: Results Negative Binomial
```{r,echo=FALSE,fig.height=3,fig.width=5,fig.align='center'}
plot(evnbinom)
```